---
title: "Lesson 6 — Survival Analysis (optional / advanced)"
subtitle: "Clinical Data Analysis in R (Phase I) — Pre-course"
author: "Neudata"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6.5, fig.height = 4.5)
```

> **This lesson is optional.** It previews the "time-to-event" analysis we cover
> later in the course. Skim it if you are curious; you do **not** need to master
> it before day 1.

## What this lesson is for

Sometimes the outcome is not just *whether* an event happened, but *when*.
**Survival analysis** handles this, including patients who were followed for
different lengths of time (**censoring**). You will meet the **Kaplan–Meier
curve**, the **log-rank test**, and **Cox regression**.

Our outcome uses two variables together:

- `time_to_event` — months of follow-up, and
- `outcome` — 1 if the event (cardiovascular event or death) occurred, 0 if the
  patient was event-free at the end (censored).

```{r}
library(tidyverse)
library(survival)    # core survival tools
library(survminer)   # nice survival plots

clinical_data <- read_csv("../Data/clinical_data_clean.csv")

# Put "Untreated" first so the hazard ratio describes the effect of being Treated
clinical_data <- clinical_data %>%
  mutate(treatment = factor(treatment, levels = c("Untreated", "Treated")))
```

## 1. The survival object

`Surv()` bundles the time and the event together.

```{r}
# Surv(time, event): time followed, and whether the event happened
surv_obj <- Surv(clinical_data$time_to_event, clinical_data$outcome)
head(surv_obj)   # a "+" means censored (no event yet)
```

## 2. Kaplan–Meier curves

**Question:** Do **treated** patients stay event-free longer than **untreated**
patients?

```{r}
# Estimate survival separately by treatment group
km_fit <- survfit(Surv(time_to_event, outcome) ~ treatment, data = clinical_data)

ggsurvplot(km_fit, data = clinical_data,
           pval = TRUE, risk.table = TRUE,
           legend.title = "Treatment",
           xlab = "Months of follow-up", ylab = "Event-free probability",
           palette = c("#C1440E", "#0D7377"))
```

**Interpretation:** each curve shows the proportion still event-free over time.
The treated curve stays higher — treated patients remain event-free longer. The
p-value (from the log-rank test) tests whether the two curves differ.

## 3. The log-rank test

```{r}
survdiff(Surv(time_to_event, outcome) ~ treatment, data = clinical_data)
```

**Interpretation:** a small p-value means the survival experience genuinely
differs between treated and untreated patients.

## 4. Cox regression — hazard ratios

The **Cox model** gives a **hazard ratio (HR)** and, like logistic regression,
lets us **adjust** for other variables.

```{r}
# Crude model: treatment only
cox_crude <- coxph(Surv(time_to_event, outcome) ~ treatment, data = clinical_data)
summary(cox_crude)
```

```{r}
# Adjusted model: account for age and comorbidity
cox_adj <- coxph(Surv(time_to_event, outcome) ~ age + diabetes + hypertension + treatment,
                 data = clinical_data)

library(broom)
tidy(cox_adj, exponentiate = TRUE, conf.int = TRUE)
```

**Output → Interpretation (hazard ratios):**

- `treatment (Treated)`: HR ≈ **0.55** — treated patients have about **45% lower**
  risk of the event at any moment, after adjustment.
- `age`: HR ≈ 1.05 per year; `diabetes`: HR ≈ 2.2; `hypertension`: HR ≈ 1.6 — all
  increase risk.

An HR **below 1** means lower risk (protective); **above 1** means higher risk.

**Reporting:** *"After adjustment for age, diabetes and hypertension, treatment
was associated with a lower hazard of the composite event (HR 0.55, 95% CI
0.36–0.85)."*

---

## Key points to remember

- Survival data = **time** + **event indicator**; `Surv(time, event)` combines them.
- **Kaplan–Meier** curves show event-free probability over time; the **log-rank**
  test compares groups.
- **Cox regression** gives adjusted **hazard ratios**; HR < 1 is protective.
- Censoring (incomplete follow-up) is handled automatically — a key advantage.

## Try it yourself (optional)

1. Draw Kaplan–Meier curves of `outcome`/`time_to_event` by `diabetes`.
2. Run a log-rank test for the diabetes comparison.
3. Fit a Cox model with `age + treatment` only and report the treatment HR.
