---
title: "Lesson 3: Exploratory Data Analysis"
subtitle: "Clinical Data Analysis in R (Phase I), Pre-course"
author: "Neudata"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 4)
```

## What this lesson is for

Before testing anything, we **describe** and **look at** the data. You will
compute descriptive statistics, build a publication-style summary table, and
make the four plots you will use most often: a histogram, a boxplot, a bar plot
and a scatterplot: and, just as importantly, **interpret** each one.

```{r}
library(tidyverse)
library(gtsummary)

clinical_data <- read_csv("../Data/clinical_data_clean.csv")

# A couple of grouped variables for plotting
clinical_data <- clinical_data %>%
  mutate(
    bmi_group = case_when(
      BMI < 18.5 ~ "Underweight",
      BMI < 25   ~ "Normal",
      BMI < 30   ~ "Overweight",
      BMI >= 30  ~ "Obese"
    ),
    bmi_group = factor(bmi_group,
                       levels = c("Underweight", "Normal", "Overweight", "Obese"))
  )
```

---

## 1. Describing continuous variables

For a continuous variable we usually report the **mean** and **standard
deviation** (if roughly symmetric) or the **median** and **IQR** (if skewed).

```{r}
mean(clinical_data$age)                 # average age
sd(clinical_data$age)                   # standard deviation
median(clinical_data$BMI)               # middle value of BMI
IQR(clinical_data$BMI)                  # interquartile range
```

## 2. Describing categorical variables

For categories we report **counts** and **percentages**.

```{r}
# Counts
table(clinical_data$smoking)

# Proportions (multiply by 100 for percentages)
prop.table(table(clinical_data$smoking)) * 100

# Prevalence of hypertension
mean(clinical_data$hypertension == "Yes") * 100
```

## 3. Summaries by group

The clinical question is usually *"does this differ between groups?"*
`group_by()` + `summarise()` answers it.

```{r}
clinical_data %>%
  group_by(sex) %>%
  summarise(
    n        = n(),
    mean_sbp = mean(systolic_bp, na.rm = TRUE),
    mean_bmi = mean(BMI, na.rm = TRUE)
  )
```

## 4. A publication-ready "Table 1"

`gtsummary` builds the baseline characteristics table journals expect, in one
command.

```{r}
clinical_data %>%
  select(age, sex, BMI, systolic_bp, diabetes, treatment, hypertension) %>%
  tbl_summary(by = hypertension) %>%
  add_p()                                # adds a p-value column
```

Read it as: for each characteristic, how the hypertensive and non-hypertensive
groups compare.

---

## 5. Four essential plots

### Histogram: the shape of one continuous variable

```{r}
ggplot(clinical_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "#0D7377", colour = "white") +
  labs(title = "Distribution of age", x = "Age (years)", y = "Number of patients")
```

**Interpretation:** age is roughly bell-shaped and centred in the mid-50s, a
typical middle-aged clinic population, with no strange spikes.

### Boxplot: a continuous variable across groups

```{r}
ggplot(clinical_data, aes(x = hypertension, y = systolic_bp, fill = hypertension)) +
  geom_boxplot() +
  labs(title = "Systolic BP by hypertension status",
       x = "Hypertension", y = "Systolic BP (mmHg)") +
  theme(legend.position = "none")
```

**Interpretation:** the box (middle 50% of patients) sits clearly higher for
hypertensive patients: as expected. The dots above the whiskers are outliers.

### Bar plot: counts of a category

```{r}
ggplot(clinical_data, aes(x = smoking, fill = smoking)) +
  geom_bar() +
  labs(title = "Number of patients by smoking status",
       x = "Smoking status", y = "Number of patients") +
  theme(legend.position = "none")
```

**Interpretation:** most patients have never smoked, with fewer former and
current smokers: useful to know before using smoking as a predictor.

### Scatterplot: two continuous variables together

```{r}
ggplot(clinical_data, aes(x = BMI, y = systolic_bp)) +
  geom_point(alpha = 0.5, colour = "#0D7377") +
  geom_smooth(method = "lm", se = TRUE, colour = "#C1440E") +
  labs(title = "Systolic BP against BMI",
       x = "BMI (kg/m2)", y = "Systolic BP (mmHg)")
```

**Interpretation:** the upward trend line suggests higher BMI is associated with
higher systolic blood pressure. A scatterplot is the right first step before a
correlation or linear regression.

---

## Key points to remember

- Continuous data → mean & SD, or median & IQR if skewed.
- Categorical data → counts and percentages (`table()`, `prop.table()`).
- `group_by()` + `summarise()` compares groups.
- `gtsummary::tbl_summary()` builds a Table 1 in one line.
- The four core plots: **histogram** (one variable's shape), **boxplot**
  (variable by group), **bar plot** (category counts), **scatterplot**
  (two continuous variables).
- Always *interpret* a plot in clinical terms, a picture is only useful if you
  can say what it means.

## Try it yourself

1. Make a histogram of `systolic_bp`. Is it roughly symmetric?
2. Make a boxplot of `BMI` by `sex`. Do men and women differ much?
3. Build a `tbl_summary()` of `age`, `BMI` and `systolic_bp` split `by = sex`.
4. Make a scatterplot of `age` (x) against `systolic_bp` (y) and describe the trend.
