---
title: "Lesson 1: R Basics"
subtitle: "Clinical Data Analysis in R (Phase I), Pre-course"
author: "Neudata"
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_knit:
    root.dir: "../../Course_Preparation/03_R_Basics"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## What this lesson is for

By the end of this short lesson you will be able to use R as a calculator,
store values in **objects**, understand **vectors** and **data frames**, and
read the course dataset into R. No previous programming experience is assumed.

> **How to work through this file:** open it in RStudio, then run each grey code
> block by clicking on it and pressing **Ctrl+Enter** (Windows) or **Cmd+Enter**
> (Mac). Watch what appears in the Console. Change a number, run it again, see
> what happens: you cannot break anything.

---

## 1. R is a calculator

The simplest thing R does is arithmetic. Type a sum and R returns the answer.

```{r}
2 + 2
140 / 90          # e.g. a blood pressure ratio
sqrt(16)          # square root
```

The `[1]` in front of the answer just means "this is the first (and only)
result".

## 2. Objects and the assignment arrow `<-`

We usually want to **store** a value and reuse it. We do that with the
assignment arrow `<-` (type it as `<` then `-`, or press **Alt+-**).

```{r}
# Store a patient's systolic blood pressure
sbp <- 152

# Now the name 'sbp' stands for 152
sbp
sbp - 12          # what it would be after treatment
```

Read `sbp <- 152` as *"sbp gets 152"*. The object now appears in the
**Environment** pane (top-right of RStudio).

## 3. Vectors: many values at once

Clinical data is not one number, it is many. A **vector** is a row of values,
made with `c()` ("combine").

```{r}
# Ages of five patients
ages <- c(54, 61, 47, 73, 58)
ages

# R works on the whole vector at once
ages + 1          # everyone one year older
mean(ages)        # average age
sd(ages)          # standard deviation
```

`mean()` and `sd()` are **functions**: you give them something inside the
brackets, they give you a result back.

## 4. Different kinds of data

R distinguishes **numbers** from **text** (called *character*) and from
**categories** (called *factors*).

```{r}
sex <- c("Female", "Male", "Female", "Female", "Male")   # text
sex

table(sex)        # count how many of each
```

`table()` is one of the most useful commands in clinical work, it counts
categories for you.

## 5. Missing values: `NA`

Real data has gaps. R writes a missing value as `NA`. Most functions need to be
told to ignore them with `na.rm = TRUE`.

```{r}
glucose <- c(5.2, 6.1, NA, 8.9, 5.5)
mean(glucose)                 # NA - because one value is missing
mean(glucose, na.rm = TRUE)   # ignores the missing value
```

## 6. Data frames: a whole dataset

A **data frame** is a table: rows are patients, columns are variables. This is
exactly how clinical data is stored. Let's read the course dataset.

```{r}
# The tidyverse gives us friendly tools for data
library(tidyverse)

# Read the clean course dataset (a CSV file) into an object
clinical_data <- read_csv("../Data/clinical_data_clean.csv")
```

Now inspect it. These five commands are the ones you will use every time you
open a new dataset.

```{r}
dim(clinical_data)     # number of rows (patients) and columns (variables)
names(clinical_data)   # the variable names
head(clinical_data)    # the first few rows
```

```{r}
# str() shows the structure: each variable and its type
str(clinical_data)
```

```{r}
# summary() gives a quick numeric summary of every variable
summary(clinical_data)
```

## 7. Getting one column with `$`

Use the dollar sign to pull out a single variable by name.

```{r}
# The average age in the dataset
mean(clinical_data$age)

# Count patients by sex
table(clinical_data$sex)

# Average systolic blood pressure
mean(clinical_data$systolic_bp)
```

---

## Key points to remember

- `<-` stores a value in a named **object**.
- `c()` builds a **vector**; functions like `mean()`, `sd()`, `table()` work on it.
- A **data frame** is a table of patients (rows) and variables (columns).
- `dim()`, `names()`, `head()`, `str()`, `summary()` inspect any dataset.
- `data$variable` selects one column.
- Missing data is `NA`; add `na.rm = TRUE` to ignore it.
- Comments start with `#`: R ignores them, humans read them.

## Try it yourself

1. Make a vector called `bp` containing the values `130, 145, 160, 138, 152`
   and calculate its mean and standard deviation.
2. Read the dataset again into `clinical_data` and use `dim()` to say how many
   patients and variables it has.
3. Use `table()` on `clinical_data$smoking`, how many current smokers are there?
4. Use `mean()` with `clinical_data$BMI`, what is the average BMI?

*Worked answers are in `Exercises/` and `Solutions/`. Don't worry if something
doesn't work first time: that is completely normal.*
