---
title: "Lesson 4 — Common Statistical Tests"
subtitle: "Clinical Data Analysis in R (Phase I) — Pre-course"
author: "Neudata"
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_knit:
    root.dir: "../../Course_Preparation/06_Statistical_Tests"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## What this lesson is for

A statistical test answers: *"could this difference be due to chance alone?"*
You will run the four tests used most in clinical research — **t-test**,
**Wilcoxon**, **chi-square / Fisher**, and **correlation** — and, crucially,
learn to **choose** and **interpret** them.

> **Choosing a test starts with the question, not the command.** Ask: what is my
> outcome, what type is it, and how many groups am I comparing? The
> `Cheat_Sheets/Statistical_Test_Decision_Guide.md` maps every situation to a test.

```{r}
library(tidyverse)
clinical_data <- read_csv("../Data/clinical_data_clean.csv")
```

---

## 1. t-test — a continuous outcome in two groups

**Question:** Is fasting **glucose** higher in patients **with** diabetes than
without?

```{r}
# Outcome: glucose (continuous). Groups: diabetes Yes/No.
t.test(glucose ~ diabetes, data = clinical_data)
```

**Interpretation:** the mean glucose is much higher in the diabetes group, and
the p-value is very small (well below 0.05), so this difference is very unlikely
to be chance. Always report the **means** and the **confidence interval for the
difference**, not just the p-value.

Now a question where the answer is *no difference*:

**Question:** Does **systolic BP** differ between men and women?

```{r}
t.test(systolic_bp ~ sex, data = clinical_data)
```

**Interpretation:** the means are almost identical and p is large (~0.85). We
conclude there is **no evidence** of a sex difference in systolic BP here. A
non-significant result is a real, reportable finding — not a failure.

## 2. Wilcoxon test — the non-parametric alternative

The t-test assumes the outcome is roughly normally distributed. If it is
skewed, use the **Wilcoxon rank-sum test** (also called Mann–Whitney), which
compares ranks instead of means.

```{r}
wilcox.test(glucose ~ diabetes, data = clinical_data)
```

**Interpretation:** it agrees with the t-test — glucose is higher with diabetes.
Use Wilcoxon when the data are skewed or have strong outliers; use the t-test
when they are approximately normal.

## 3. Chi-square test — two categorical variables

**Question:** Is **hypertension** associated with **diabetes**?

```{r}
# A cross-tabulation first
table(clinical_data$hypertension, clinical_data$diabetes)

# Then the test
chisq.test(table(clinical_data$hypertension, clinical_data$diabetes))
```

**Interpretation:** the small p-value (< 0.001) means hypertension and diabetes
occur together more than expected by chance — they are associated. Chi-square
assumes expected counts are mostly 5 or more.

## 4. Fisher's exact test — when counts are small

When any expected count is small, use **Fisher's exact test** instead.

```{r}
# Example on a small sub-table (smokers only)
smokers <- clinical_data %>% filter(smoking == "Current")
fisher.test(table(smokers$hypertension, smokers$diabetes))
```

**Interpretation:** Fisher gives a valid p-value even in small samples. For
2×2 tables it also returns an odds ratio.

## 5. Correlation — two continuous variables

**Question:** Do **age** and **systolic BP** move together?

```{r}
cor.test(clinical_data$age, clinical_data$systolic_bp)
```

**Interpretation:** the correlation coefficient (r ≈ 0.55) is positive and
moderate — older patients tend to have higher systolic BP — and the p-value is
tiny. Use **Pearson** correlation for straight-line relationships and
**Spearman** (`method = "spearman"`) if the relationship is non-linear or the
data are skewed.

---

## Reading any test result

Every test above prints:

- an **estimate** (a difference in means, an odds ratio, a correlation) — *the
  clinically important number*;
- a **confidence interval** — the range of plausible values;
- a **p-value** — how compatible the data are with "no effect".

Report the **estimate and its confidence interval first**; the p-value supports
them. A p-value alone never tells you how *big* or *important* an effect is.

## Key points to remember

- Match the test to the question: outcome type + number of groups (see the
  decision guide).
- **t-test / Wilcoxon** — continuous outcome, two groups.
- **chi-square / Fisher** — two categorical variables (Fisher for small counts).
- **correlation** — two continuous variables.
- Non-significant is still a finding.
- Always report the effect estimate and confidence interval, not just *p*.

## Try it yourself

1. Use a t-test to compare **BMI** between hypertensive and non-hypertensive
   patients. Is the difference significant?
2. Cross-tabulate `sex` and `treatment` with `table()`, then test the
   association with `chisq.test()`.
3. Test the correlation between `BMI` and `systolic_bp`. Is it stronger or
   weaker than age vs systolic BP?
4. For each of the three questions above, write one sentence stating the estimate
   *and* the p-value.
