---
title: "Lesson 2 — Data Management (Cleaning Clinical Data)"
subtitle: "Clinical Data Analysis in R (Phase I) — Pre-course"
author: "Neudata"
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_knit:
    root.dir: "../../Course_Preparation/04_Data_Management"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## What this lesson is for

Real clinical data is messy. Before any analysis we must **clean** it. In this
lesson you will find and fix the common problems — duplicates, missing values,
inconsistent labels, and impossible values — using a small set of `dplyr` verbs:
`filter()`, `select()`, `mutate()`, `case_when()`, `group_by()` and
`summarise()`.

We will work on the **raw** file (`clinical_data_raw.csv`), which contains
deliberate problems described in `Data/Data_Quality_Problems.md`.

```{r}
library(tidyverse)

# Read the messy raw dataset
raw <- read_csv("../Data/clinical_data_raw.csv")

dim(raw)      # note: more rows than patients, because of duplicates
```

---

## 1. Look before you leap

Always inspect a new dataset first.

```{r}
# How many missing values in each column?
colSums(is.na(raw))

# How is 'sex' recorded? (look for inconsistent labels)
table(raw$sex)

# Quick numeric overview - watch for impossible minimums/maximums
summary(select(raw, age, height_cm, weight_kg, BMI, systolic_bp, glucose, cholesterol))
```

You should see: several missing values, `sex` spelled many different ways, and
some impossible numbers (an age of 219, a height of 17 cm, a glucose of 41).

## 2. Remove duplicate records

```{r}
# How many exact duplicate rows?
sum(duplicated(raw))

# Keep only unique rows
clean <- raw %>% distinct()

nrow(clean)   # should now be 430 unique patients
```

## 3. Make categories consistent with `case_when()`

`case_when()` reads like a set of clinical rules: *if this, then that*.

```{r}
clean <- clean %>%
  mutate(
    # Collapse all the spellings of sex into two clean labels
    sex = case_when(
      sex %in% c("Female", "female", "F", "f") ~ "Female",
      sex %in% c("Male", "male", "M", "m")     ~ "Male"
    ),
    # Standardise Yes/No variables (they were mixed with 1/0 and yes/no)
    hypertension = case_when(
      hypertension %in% c("Yes", "yes", "1") ~ "Yes",
      hypertension %in% c("No",  "no",  "0") ~ "No"
    ),
    diabetes = case_when(
      diabetes %in% c("Yes", "yes", "1") ~ "Yes",
      diabetes %in% c("No",  "no",  "0") ~ "No"
    ),
    # Fix the stray lower-case "current"
    smoking = case_when(
      smoking %in% c("Current", "current") ~ "Current",
      TRUE ~ smoking      # keep Never / Former unchanged
    )
  )

table(clean$sex)
table(clean$hypertension)
```

## 4. Deal with impossible values

We cannot know the true value of an impossible entry, so we set it to `NA`.
`if_else()` says: *if the condition is true, use the first value, otherwise the
second.*

```{r}
clean <- clean %>%
  mutate(
    age         = if_else(age < 18 | age > 110, NA_real_, age),
    height_cm   = if_else(height_cm < 100 | height_cm > 220, NA_real_, height_cm),
    weight_kg   = if_else(weight_kg < 30 | weight_kg > 250, NA_real_, weight_kg),
    systolic_bp = if_else(systolic_bp > 260, NA_real_, systolic_bp),
    glucose     = if_else(glucose > 30, NA_real_, glucose),      # impossible fasting glucose
    cholesterol = if_else(cholesterol < 1, NA_real_, cholesterol) # negative lab value
  )

# Recompute BMI from the cleaned weight and height so it is consistent
clean <- clean %>%
  mutate(BMI = round(weight_kg / (height_cm / 100)^2, 1))

summary(select(clean, age, height_cm, weight_kg, BMI, systolic_bp, glucose))
```

The impossible minimums and maximums are gone.

## 5. Fix the date inconsistency

```{r}
# A follow-up date should never be before the admission date
sum(clean$followup_date < clean$admission_date, na.rm = TRUE)

clean <- clean %>%
  mutate(followup_date = if_else(followup_date < admission_date,
                                 as.Date(NA), followup_date))
```

## 6. Create new (derived) variables

Analysis often needs grouped or calculated variables. Here we build three.

```{r}
clean <- clean %>%
  mutate(
    # Clinical BMI categories
    bmi_group = case_when(
      BMI < 18.5             ~ "Underweight",
      BMI < 25               ~ "Normal",
      BMI < 30               ~ "Overweight",
      BMI >= 30              ~ "Obese"
      # note: leaving out a "TRUE ~" line means missing BMI stays NA
    ),
    # Age groups
    age_group = case_when(
      age < 40               ~ "<40",
      age < 55               ~ "40-54",
      age < 70               ~ "55-69",
      age >= 70              ~ "70+"
    ),
    # A derived clinical measure: pulse pressure
    pulse_pressure = systolic_bp - diastolic_bp
  )

table(clean$bmi_group)
table(clean$age_group)
```

## 7. The core `dplyr` verbs

These six verbs cover most day-to-day data work.

```{r}
# filter(): keep rows that meet a condition
hypertensive <- clean %>% filter(hypertension == "Yes")
nrow(hypertensive)

# select(): keep only certain columns
small <- clean %>% select(patient_id, age, sex, BMI, hypertension)
head(small)

# group_by() + summarise(): summary statistics by group
clean %>%
  group_by(sex) %>%
  summarise(
    n           = n(),
    mean_age    = mean(age, na.rm = TRUE),
    mean_sbp    = mean(systolic_bp, na.rm = TRUE)
  )
```

## 8. Save the analysis-ready dataset

```{r, eval=FALSE}
# Write the cleaned data so later lessons can use it
write_csv(clean, "../Data/my_clean_data.csv")
```

> A ready-made reference version is already provided as
> `Data/clinical_data_clean.csv`. From Lesson 3 onward we simply load that file,
> so everyone starts from exactly the same point.

---

## Key points to remember

- Always **inspect** first: `colSums(is.na())`, `table()`, `summary()`.
- `distinct()` removes duplicate rows.
- `case_when()` standardises messy categories using clinical rules.
- Set **impossible values** to `NA` — never invent a value.
- Recompute derived variables (like `BMI`) *after* fixing their inputs.
- `filter()`, `select()`, `mutate()`, `group_by()`, `summarise()` are the
  everyday verbs of data management.

## Try it yourself

1. After cleaning, how many patients are **current smokers**? (`table()`)
2. Use `filter()` to keep only patients aged 60 or over. How many are there?
3. Create a variable `high_bp` that is `"High"` when `systolic_bp >= 140` and
   `"Normal"` otherwise, using `if_else()`.
4. Use `group_by(hypertension)` and `summarise()` to compare mean BMI in
   hypertensive vs non-hypertensive patients.
