---
title: "Lesson 5: Regression"
subtitle: "Clinical Data Analysis in R (Phase I), Pre-course"
author: "Neudata"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## What this lesson is for

Regression lets us study several predictors at once and **adjust** for
confounders. You will fit and interpret the two workhorses of clinical research:
**linear regression** (for a continuous outcome) and **logistic regression**
(for a yes/no outcome), and turn the output into a sentence you could put in a
paper.

We write models plainly, so you can see exactly what is being estimated:

```r
model <- glm(hypertension ~ age + sex + BMI,
             data = clinical_data,
             family = binomial)
```

```{r}
library(tidyverse)
library(broom)       # tidies model output
library(gtsummary)   # publication-ready model tables

clinical_data <- read_csv("../Data/clinical_data_clean.csv")
```

---

## 1. Linear regression: a continuous outcome

**Question:** Is **systolic BP** associated with **age** and **BMI**?

**Data:** outcome `systolic_bp` (continuous); predictors `age`, `BMI` (continuous).

```{r}
# lm() = linear model
lin_model <- lm(systolic_bp ~ age + BMI, data = clinical_data)
summary(lin_model)
```

**Output → Interpretation:** each coefficient is the change in systolic BP per
one-unit increase in that predictor, holding the other constant.

- `age`: about +0.35 mmHg for each additional year.
- `BMI`: about +0.6 mmHg for each additional BMI unit.
- Both p-values are very small, so both associations are statistically
  significant.

```{r}
# Tidy estimates with 95% confidence intervals
tidy(lin_model, conf.int = TRUE)
```

**Reporting:** *"Each additional year of age was associated with a 0.35 mmHg
higher systolic BP (95% CI ...), and each BMI unit with a 0.62 mmHg increase,
after mutual adjustment."*

## 2. Logistic regression: a yes/no outcome

When the outcome is binary (Yes/No), we use **logistic regression**, which gives
**odds ratios (OR)**.

**Question:** What is associated with **having hypertension**?

**Data:** outcome `hypertension` (Yes/No); predictors `age`, `sex`, `BMI`.

```{r}
# Make hypertension a factor with "No" as the reference level
clinical_data <- clinical_data %>%
  mutate(hypertension = factor(hypertension, levels = c("No", "Yes")))

# glm() with family = binomial fits a logistic regression
log_model <- glm(hypertension ~ age + sex + BMI,
                 data = clinical_data,
                 family = binomial)

summary(log_model)
```

The `summary()` output is on the **log-odds** scale, hard to read. We convert to
**odds ratios** by exponentiating.

```{r}
# exponentiate = TRUE turns log-odds into odds ratios
tidy(log_model, exponentiate = TRUE, conf.int = TRUE)
```

**Output → Interpretation (odds ratios):**

- `age`: OR ≈ **1.09**: each extra year raises the odds of hypertension by ~9%.
- `BMI`: OR ≈ **1.16**: each extra BMI unit raises the odds by ~16%.
- `sex (Male)`: OR ≈ **1.20**, but the confidence interval crosses 1 and p ≈ 0.43
, **no evidence** of a sex difference.

An OR **above 1** means higher odds; **below 1** means lower odds; a CI that
**includes 1** means "not statistically significant".

## 3. A publication-ready regression table

```{r}
tbl_regression(log_model, exponentiate = TRUE)
```

This is the table format journals expect: each predictor with its OR, 95% CI and
p-value.

## 4. Adjustment and confounding (why regression matters)

Older patients tend to have higher BMI *and* more hypertension. A simple
comparison could mistake age's effect for BMI's. Putting both in the same model
**adjusts** each for the other, so each odds ratio is "independent of" the other
predictors. This is the core reason clinical research uses multivariable models.

---

## Key points to remember

- **Linear regression** (`lm`): continuous outcome; coefficients are changes in
  the outcome per unit of predictor.
- **Logistic regression** (`glm(..., family = binomial)`), binary outcome;
  exponentiated coefficients are **odds ratios**.
- `tidy(model, exponentiate = TRUE, conf.int = TRUE)` gives OR + 95% CI.
- OR > 1 raises the odds, < 1 lowers it; a CI crossing 1 is non-significant.
- Multivariable models **adjust** for confounders, the main reason we use them.
- Report the estimate and its confidence interval in words, never raw output.

## Try it yourself

1. Fit `lm(diastolic_bp ~ age + BMI)`. How much does diastolic BP change per BMI
   unit?
2. Fit a logistic model `diabetes ~ age + BMI`. Report the odds ratios.
3. Add `diabetes` to the hypertension model (`hypertension ~ age + sex + BMI +
   diabetes`). Does the OR for diabetes suggest an association?
4. Write one publication-style sentence reporting the age and BMI odds ratios
   from the hypertension model.
