# Regression Modelling and Reporting Results {#ch-regression}

In Chapter 4 we asked one question at a time. Is treatment uptake more
common among patients with health insurance? Do urban and rural patients
differ? Each test compared groups defined by a single variable. Clinical
reality is rarely that tidy. Patients with health insurance may also be
younger, better educated or live closer to a clinic, and each of these
characteristics may itself influence whether a person with hypertension
starts treatment. If we want to know whether insurance is associated with
uptake *over and above* these other factors, we need a method that
considers several variables at once. That method is regression.

Regression is the workhorse of clinical and epidemiological research. It
appears in almost every observational study published in medical journals,
and it is the natural end point of the analysis we have been building since
Chapter 1: we imported and cleaned the hypertension data, described the
sample, tested simple hypotheses, and now we bring the determinants of
treatment uptake together in a single model. This chapter explains how
regression models work, how to fit and read them in R, how to check that
they are reasonable, and how to turn the output into the tables, figures
and sentences that appear in a published paper. It finishes with the
practices that make the whole analysis reproducible, so that another
person (or you, a year from now) can obtain exactly the same results.

::: {.objectives}
- Explain the three purposes of regression modelling (description,
  prediction and explanation) and why the purpose shapes the model.
- Fit and interpret a multiple linear regression for a continuous outcome,
  including coefficients, $R^2$ and residual checks.
- Convert between probabilities, odds and log-odds, and explain why
  logistic regression models the log-odds of a binary outcome.
- Fit a multivariable logistic regression with `glm()`, read its
  `summary()` output line by line, and report odds ratios with 95%
  confidence intervals.
- Distinguish crude from adjusted estimates, recognise confounding, and
  avoid the "Table 2 fallacy".
- Build a model from clinical reasoning rather than automated selection,
  and check events per variable, collinearity, linearity, discrimination,
  calibration and influential observations.
- Produce a publication-ready regression table and a forest plot, and write
  a Results paragraph that follows the STROBE recommendations.
- Organise an analysis so that it can be reproduced from raw data to report.
:::

## Why regression? {#sec-why-regression}

### From one variable at a time to many

The tests in Chapter 4 answer questions of the form "is the outcome
associated with *this* variable?". A regression model answers a richer
question: "how is the outcome associated with this variable, *given* the
values of these other variables?" It does this by writing the expected
value of the outcome (or a transformation of it) as a function of several
predictors simultaneously. Each coefficient in the model then describes the
association between one predictor and the outcome while the other
predictors are held fixed. This "holding fixed" is what epidemiologists
call *adjustment*, and it is the main reason regression is used in
observational research [@vittinghoff2012; @kirkwood2003].

Regression also provides more than a p-value. It gives an *estimate* of the
size of each association, with a confidence interval describing its
precision, and it can produce predictions for individual patients. As
Chapter 4 emphasised, an estimate with its interval is far more informative
than a statement that a test was "significant" [@altman1991;
@wasserstein2016].

### Description, prediction and explanation

Before fitting any model you should be clear about what it is *for*,
because the purpose determines which variables to include and how to judge
the result [@vittinghoff2012; @harrell2015].

- **Description.** The aim is to summarise how an outcome varies with
  several characteristics in a population, for example how the prevalence
  of treatment uptake differs by age, sex and residence. The model is a
  compact, multivariable summary; no causal claim is made.
- **Prediction.** The aim is to estimate, as accurately as possible, the
  outcome for a new individual, for instance the probability that a newly
  diagnosed patient will start treatment. Any variable that improves
  predictive accuracy is welcome, whether or not it is a cause. Models are
  judged by discrimination and calibration in *new* data, and great care is
  needed to avoid overfitting [@harrell2015].
- **Explanation (aetiology).** The aim is to estimate the effect of a
  particular exposure on the outcome, for example the effect of health
  insurance on uptake. Variables are included because they are
  *confounders* of that exposure–outcome relationship, and the choice is
  guided by subject-matter knowledge, often summarised in a causal diagram
  [@rothman2008; @greenland1999dag].

Our case study, *Determinants of hypertension treatment uptake among adults
attending primary healthcare facilities*, sits between description and
explanation. We want to know which patient and access characteristics are
independently associated with uptake, so that services can be targeted.
This chapter therefore fits a single, pre-specified multivariable model and
interprets its coefficients as *adjusted associations*. We shall return
repeatedly to the danger of reading them as causal effects.

::: {.callout-tip title="Good practice"}
Write down the purpose of your model, and the list of variables you will
include, in your analysis plan *before* you look at the outcome data. A
prediction model, a descriptive model and an aetiological model built from
the same data set can legitimately look quite different.
:::

### A family of regression models

All the models in this chapter share one structure: a *linear predictor*,
which is a weighted sum of the explanatory variables,

$$
\eta = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_k x_k ,
$$

linked to the expected value of the outcome by a function chosen to suit the
type of outcome. These are the *generalised linear models*
[@agresti2013]. The type of outcome determines the model, as summarised
below.

Table: Choosing a regression model by the type of outcome.

| Outcome type | Example in the study | Model | Effect measure | R function |
|---|---|---|---|---|
| Continuous | Systolic blood pressure | Linear | Difference in means | `lm()` |
| Binary | Treatment uptake (Yes/No) | Logistic | Odds ratio | `glm(family = binomial)` |
| Count | Number of comorbidities | Poisson | Rate ratio | `glm(family = poisson)` |
| Time to event | Time to treatment start | Cox | Hazard ratio | `survival::coxph()` |

This chapter covers the first two rows in detail. Once you understand linear
and logistic regression, the others follow the same logic.

## Setting up the analysis {#sec-setup}

As in every chapter, we start a fresh R session by loading the packages and
the clean analysis data set created in Chapter 2.

```{r c5-setup}
library(tidyverse)   # dplyr, ggplot2, tidyr, ...
library(broom)       # tidy() and augment() for model objects
library(gtsummary)   # publication-ready regression tables
library(car)         # vif() for collinearity
library(pROC)        # ROC curves and the AUC
library(patchwork)   # combining ggplot2 panels
library(scales)      # axis formatting
library(knitr)       # kable() tables

analysis_data <- readRDS("Data/analysis_data.rds")
```

The primary outcome, `treatment_uptake`, only makes sense for people who
already know they have hypertension: a person who has never been diagnosed
cannot "take up" treatment. The analytic population is therefore the
patients with `htn_diagnosed == "Yes"`.

```{r c5-population}
dx <- analysis_data |>
  filter(htn_diagnosed == "Yes") |>
  # education is stored as an ORDERED factor (None < Primary < ...).
  # For regression we convert it to an ordinary factor so that each level
  # is compared with the reference level "None" (explained below).
  mutate(education = factor(education, ordered = FALSE))

nrow(dx)                         # size of the analytic population
table(dx$treatment_uptake)       # the outcome distribution
levels(dx$treatment_uptake)      # first level = reference ("No")
```

Of the 1,089 diagnosed patients, 508 (46.6%) are on antihypertensive
treatment. The first level of the outcome factor is `"No"`, so the models
below estimate the odds of *uptake* ("Yes"), which is the event of
interest. If the levels were the other way round, every odds ratio in this
chapter would be inverted.

::: {.callout-warning title="Common mistake"}
Ordered factors are convenient for tables and plots, but R codes them in
regression with *orthogonal polynomial contrasts*. The output then contains
terms called `education.L`, `education.Q` and `education.C` (linear,
quadratic and cubic trends) instead of one comparison per level. Their odds
ratios are not "Primary versus None" or "per level of education", and they
are very easy to misreport. Unless you specifically want a trend test,
convert ordered factors with `factor(x, ordered = FALSE)` before modelling,
as we did above.
:::

## Linear regression for a continuous outcome {#sec-linear}

Linear regression is the simplest regression model and the foundation for
everything else, so we begin with it, using systolic blood pressure (SBP)
as the outcome. Because SBP is measured in everyone, we use the full sample
of 1,500 adults for this section.

### The simple linear regression model

With one predictor $x$ (say body mass index, BMI), the simple linear
regression model for patient $i$ is

$$
y_i = \beta_0 + \beta_1 x_i + \varepsilon_i ,
\qquad \varepsilon_i \sim N(0, \sigma^2) .
$$

Here $\beta_0$ is the *intercept* (the mean of $y$ when $x = 0$), $\beta_1$
is the *slope* (the change in the mean of $y$ for a one-unit increase in
$x$), and $\varepsilon_i$ is the random error: the part of patient $i$'s
blood pressure not explained by BMI. The errors are assumed to be
independent and normally distributed with constant variance $\sigma^2$.

The coefficients are estimated by *ordinary least squares*: we choose
$\hat\beta_0$ and $\hat\beta_1$ to make the sum of squared vertical
distances between the observed points and the fitted line as small as
possible,

$$
\text{minimise} \quad \sum_{i=1}^{n} \left(y_i - \beta_0 - \beta_1 x_i\right)^2 ,
$$

which gives the closed-form solution

$$
\hat\beta_1 = \frac{\sum_{i}(x_i - \bar x)(y_i - \bar y)}{\sum_{i}(x_i - \bar x)^2},
\qquad
\hat\beta_0 = \bar y - \hat\beta_1 \bar x .
$$

The slope is therefore closely related to the Pearson correlation of
Chapter 4: $\hat\beta_1 = r \, s_y / s_x$ [@altman1991; @bland2015].

```{r c5-lm-simple}
# Simple linear regression: SBP (mmHg) on BMI (kg/m^2)
lm_simple <- lm(sbp_mmhg ~ bmi, data = analysis_data)
coef(lm_simple)
```

The fitted line is $\widehat{\text{SBP}} =
`r round(coef(lm_simple)[1], 1)` + `r round(coef(lm_simple)[2], 2)`
\times \text{BMI}$. Each additional kg/m² of BMI is associated with a mean
SBP about `r round(coef(lm_simple)[2], 2)` mmHg higher. The intercept is the
predicted SBP at a BMI of zero, which is biologically impossible; it is
needed to position the line but has no clinical interpretation. This is
common, and it is one reason some analysts *centre* continuous predictors
(subtract a typical value such as the mean) before fitting.

### Multiple linear regression

Blood pressure rises with age, and BMI may also differ by age, so we add age
to the model. With $k$ predictors the model becomes

$$
y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \dots + \beta_k x_{ki}
+ \varepsilon_i ,
$$

or, in matrix notation, $\mathbf{y} = \mathbf{X}\boldsymbol\beta +
\boldsymbol\varepsilon$, with least-squares solution
$\hat{\boldsymbol\beta} = (\mathbf{X}^\top\mathbf{X})^{-1}
\mathbf{X}^\top\mathbf{y}$. You will never compute this by hand, but the
formula makes one point clear: each coefficient is estimated using all the
predictors together, so its value depends on which other variables are in
the model. Each $\beta_j$ is the expected difference in $y$ between two
people who differ by one unit in $x_j$ but have *the same values* of all
the other predictors [@vittinghoff2012].

```{r c5-lm-multiple}
lm_sbp <- lm(sbp_mmhg ~ age + bmi, data = analysis_data)
summary(lm_sbp)
```

Let us read this output from top to bottom.

- **Call** repeats the model formula and data, which is useful when you
  have fitted many models.
- **Residuals** summarises the differences between observed and fitted
  SBP. The median is close to zero and the quartiles are roughly symmetric
  (about $\pm 12.6$ mmHg), which is what we expect if the errors are
  symmetric around zero.
- **Coefficients** is the heart of the output. For each term there is the
  estimate, its standard error (SE), the $t$ statistic (estimate divided by
  SE) and the two-sided p-value for the null hypothesis that the true
  coefficient is zero.
  - `(Intercept)` 87.5 mmHg is the predicted SBP for a person aged 0 with
    a BMI of 0, again not clinically meaningful.
  - `age` 0.26: holding BMI fixed, each additional year of age is
    associated with a mean SBP 0.26 mmHg higher; over ten years that is
    about 2.6 mmHg.
  - `bmi` 1.45: among people of the same age, each additional kg/m² is
    associated with a mean SBP 1.45 mmHg higher; a 5 kg/m² difference
    corresponds to about 7.3 mmHg.
- **Residual standard error** (18.1 mmHg) estimates $\sigma$, the typical
  distance of an observation from its fitted value. It is large relative to
  the coefficients: age and BMI explain only part of the variation in blood
  pressure.
- **`(51 observations deleted due to missingness)`** reminds us that `lm()`
  silently drops every row with a missing value in any model variable. The
  model was fitted to 1,449 people, not 1,500.
- **Multiple R-squared** (0.16) is the proportion of the variance of SBP
  explained by the model,
  $R^2 = 1 - \sum_i (y_i - \hat y_i)^2 / \sum_i (y_i - \bar y)^2$. The
  **adjusted** $R^2$ applies a small penalty for the number of predictors.
- **F-statistic** tests the null hypothesis that *all* the slope
  coefficients are zero. With $p < 2.2 \times 10^{-16}$ we conclude that age
  and BMI together explain some variation in SBP.

::: {.callout-note title="Clinical interpretation"}
An $R^2$ of 0.16 does not mean the model is useless. In an explanatory
analysis we care about the *coefficients*: a difference of about 7 mmHg in
mean SBP for 5 kg/m² of BMI, independent of age, is clinically important at
the population level. A low $R^2$ simply tells us that individual blood
pressure depends on many other things (genetics, salt intake, medication,
measurement error) that are not in the model.
:::

To make the coefficients easier to quote, we can ask for confidence
intervals and rescale age to a more meaningful unit.

```{r c5-lm-ci}
tidy(lm_sbp, conf.int = TRUE) |>
  select(term, estimate, conf.low, conf.high, p.value) |>
  mutate(across(estimate:conf.high, \(x) round(x, 3)),
         p.value = signif(p.value, 2))

# Per 10 years of age: multiply the age coefficient and its CI by 10
age10 <- 10 * c(coef(lm_sbp)["age"], confint(lm_sbp)["age", ])
round(age10, 2)
```

We would report: "Each 10-year increase in age was associated with a
`r sprintf("%.1f", age10[1])` mmHg higher mean SBP (95% CI
`r sprintf("%.1f", age10[2])` to `r sprintf("%.1f", age10[3])`), adjusted
for BMI."

### Assumptions and residual checks

The inferences from a linear model (standard errors, confidence intervals,
p-values) rely on four assumptions, conveniently remembered as **LINE**
[@vittinghoff2012; @kirkwood2003]:

1. **L**inearity: the mean of $y$ changes linearly with each continuous
   predictor (given the others).
2. **I**ndependence: the errors of different patients are independent.
   This is a property of the study design; repeated measurements on the same
   patient, or patients clustered within facilities, can violate it.
3. **N**ormality: the errors are approximately normally distributed. With
   large samples the central limit theorem makes the coefficients
   approximately normal even if the errors are not, so this matters most in
   small samples and for prediction intervals.
4. **E**qual variance (homoscedasticity): the spread of the errors does not
   change with the fitted value.

We check these assumptions with the *residuals*,
$e_i = y_i - \hat y_i$. Two plots are standard: residuals against fitted
values (for linearity and equal variance) and a normal Q–Q plot of the
residuals (for normality). `broom::augment()` returns the fitted values and
residuals in a data frame that is convenient for `ggplot2`.

```{r c5-lm-residuals, fig.cap="Residual diagnostics for the linear regression of systolic blood pressure on age and BMI. Left: residuals against fitted values, with a smooth curve; right: normal Q-Q plot of the residuals.", fig.height=3.6}
lm_aug <- augment(lm_sbp)      # one row per patient used in the model

p_resfit <- ggplot(lm_aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.25, colour = "#0D7377") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, colour = "#E07A1F") +
  labs(title = "Residuals vs fitted",
       x = "Fitted SBP (mmHg)", y = "Residual (mmHg)")

p_qq <- ggplot(lm_aug, aes(sample = .std.resid)) +
  stat_qq(alpha = 0.25, colour = "#0D7377") +
  stat_qq_line(linetype = "dashed") +
  labs(title = "Normal Q-Q plot",
       x = "Theoretical quantiles", y = "Standardised residual")

p_resfit + p_qq
```

In the left panel the points form a horizontal band of roughly constant
width centred on zero, and the smooth curve is close to flat apart from a
slight upturn at the highest fitted values, where there are few patients
and the smoother is unstable. There is no convincing sign of curvature
(which would suggest a missing non-linear term) or of a
funnel shape (which would suggest unequal variance). In the right panel the
points follow the reference line closely, with only slight departures in
the extreme tails. The assumptions look reasonable.

::: {.callout-tip title="Good practice"}
Look at residual plots, do not test them. Formal tests of normality reject
trivial departures in large samples and miss important ones in small
samples [@altman1995normal]. If a plot shows a curved pattern, consider
adding a non-linear term (for example a quadratic or spline in age); if it
shows a funnel, consider transforming the outcome or using robust standard
errors.
:::

## Binary outcomes: probability, odds and log-odds {#sec-odds}

Our primary outcome, treatment uptake, is binary: each patient is either on
treatment ($y = 1$) or not ($y = 0$). Linear regression is not appropriate
here. A straight line in $x$ can predict "probabilities" below 0 or above 1,
the variance of a binary variable depends on its mean (so the equal
variance assumption fails), and the errors cannot be normal. We need a
scale on which a linear model makes sense. That scale is the log-odds.

### Three ways to express the same risk

Let $p$ be the probability that a patient takes up treatment. The *odds*
of uptake are the probability of uptake divided by the probability of no
uptake,

$$
\text{odds} = \frac{p}{1-p},
\qquad\text{and conversely}\qquad
p = \frac{\text{odds}}{1+\text{odds}} .
$$

Odds are familiar from betting: odds of 3 ("3 to 1") mean the event is
three times as likely to happen as not, which is a probability of
$3/4 = 0.75$. While a probability is confined to $[0, 1]$, odds range from
0 to infinity. Taking the natural logarithm gives the *log-odds* or
*logit*,

$$
\operatorname{logit}(p) = \log\!\left(\frac{p}{1-p}\right),
$$

which ranges over the whole real line, from $-\infty$ to $+\infty$, and is
symmetric around zero: a probability of 0.5 has odds 1 and log-odds 0.

Table: The same risk expressed as a probability, odds and log-odds.

| Probability $p$ | Odds $p/(1-p)$ | Log-odds | Interpretation |
|---|---|---|---|
| 0.10 | 0.11 | $-2.20$ | 1 success for every 9 failures |
| 0.25 | 0.33 | $-1.10$ | 1 for every 3 |
| 0.50 | 1.00 | 0.00 | Even chances |
| 0.75 | 3.00 | 1.10 | 3 for every 1 |
| 0.90 | 9.00 | 2.20 | 9 for every 1 |

### The odds ratio

To compare two groups we take the ratio of their odds. If $p_1$ is the
probability of uptake among insured patients and $p_0$ among uninsured
patients, the *odds ratio* (OR) is

$$
\text{OR} = \frac{p_1/(1-p_1)}{p_0/(1-p_0)} .
$$

An OR of 1 means no association; above 1, higher odds of the outcome in
the first group; below 1, lower odds [@bland2000oddsratio]. Let us compute
it by hand for health insurance.

```{r c5-or-by-hand}
tab_ins <- table(Insurance = dx$health_insurance,
                 Uptake    = dx$treatment_uptake)
tab_ins

# Probability, odds and log-odds of uptake in each insurance group
ins_summary <- dx |>
  group_by(health_insurance) |>
  summarise(n = n(),
            p_uptake = mean(treatment_uptake == "Yes")) |>
  mutate(odds = p_uptake / (1 - p_uptake),
         log_odds = log(odds))
kable(ins_summary, digits = 3,
      caption = "Uptake by health insurance status (diagnosed patients).")

# The crude odds ratio: odds among insured / odds among uninsured
or_ins <- ins_summary$odds[2] / ins_summary$odds[1]
round(or_ins, 2)
```

```{r c5-ins-values, include=FALSE}
p0_ins <- ins_summary$p_uptake[1]     # risk of uptake, uninsured
p1_ins <- ins_summary$p_uptake[2]     # risk of uptake, insured
rr_ins <- p1_ins / p0_ins             # relative risk (risk ratio)
```

Among uninsured patients `r round(100 * p0_ins, 1)`% are on treatment
(odds `r round(ins_summary$odds[1], 2)`); among insured patients
`r round(100 * p1_ins, 1)`% (odds `r round(ins_summary$odds[2], 2)`). The
odds of uptake are `r round(or_ins, 2)` times higher among the insured.
Notice the last column of the table: the *difference* in log-odds,
`r round(ins_summary$log_odds[2], 3)` $-$
(`r round(ins_summary$log_odds[1], 3)`) $=$ `r round(log(or_ins), 3)`, is
the log of the odds ratio. This is the key to logistic regression: a
difference on the log-odds scale is a log odds ratio, and exponentiating it
gives the odds ratio.

### Odds ratios versus relative risks

Clinicians usually think in terms of risks, and the *relative risk*
(risk ratio, RR) $p_1 / p_0$ is the more intuitive measure. Here the RR is
`r round(p1_ins, 3)` / `r round(p0_ins, 3)` $=$ `r round(rr_ins, 2)`,
noticeably smaller than the OR of `r round(or_ins, 2)`. The two measures
are close only when the outcome is rare (say, below 10%) in both groups,
because then $1 - p \approx 1$ and the odds approximate the risk
[@bland2000oddsratio; @rothman2008]. When the outcome is common, as here
(almost half of patients are on treatment), the OR is always further from
1 than the RR.

@zhang1998 proposed a simple conversion from an OR to an approximate RR,
given the risk $p_0$ in the reference group:

$$
\text{RR} \approx \frac{\text{OR}}{(1 - p_0) + p_0 \times \text{OR}} .
$$

With $p_0 = `r round(p0_ins, 3)`$ and OR $= `r round(or_ins, 2)`$, the
formula returns `r round(or_ins / ((1 - p0_ins) + p0_ins * or_ins), 2)`,
which reproduces the crude RR exactly. For adjusted odds ratios the
conversion is only approximate, and other methods (such as log-binomial or
Poisson regression with robust standard errors) are preferred if an
adjusted RR is the target.

::: {.callout-warning title="Common mistake"}
Do not describe an odds ratio of 2.0 as "twice the risk" or "twice as
likely" when the outcome is common. With an uptake of about 40% in the
reference group, an OR of 2.0 corresponds to a risk ratio of only about 1.4.
Say "twice the *odds*", and, if your readers need risks, report predicted
probabilities or a risk ratio as well.
:::

Why, then, do we use odds ratios at all? There are three good reasons.
First, the logistic model on which they are based always produces valid
probabilities between 0 and 1. Second, the OR is symmetric: the OR for
uptake is the reciprocal of the OR for non-uptake, which is not true of the
RR. Third, the OR can be estimated from case–control studies, where risks
cannot [@rothman2008; @hosmer2013].

## The logistic regression model {#sec-logistic-model}

### The model equation

Logistic regression models the log-odds of the outcome as a linear
function of the predictors [@hosmer2013; @agresti2013]:

$$
\operatorname{logit}(p_i) =
\log\!\left(\frac{p_i}{1-p_i}\right) =
\beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + \dots + \beta_k x_{ki},
$$

where $p_i = \Pr(y_i = 1 \mid x_{1i}, \dots, x_{ki})$ is the probability of
uptake for patient $i$. Solving for $p_i$ gives the *logistic function*,

$$
p_i = \frac{\exp(\beta_0 + \beta_1 x_{1i} + \dots + \beta_k x_{ki})}
{1 + \exp(\beta_0 + \beta_1 x_{1i} + \dots + \beta_k x_{ki})} ,
$$

an S-shaped curve that always lies between 0 and 1. There is no error term
in the equation: the randomness lies in the outcome itself, which is
assumed to follow a Bernoulli distribution with probability $p_i$.

The figure below shows the two faces of the model for a single predictor, age.
On the log-odds scale (left) the relationship is a straight line; on the
probability scale (right) the same model is a gentle S-shaped curve.

```{r c5-logit-curve, fig.cap="A logistic regression of treatment uptake on age, shown on the log-odds scale (left), where the model is linear, and on the probability scale (right), where it is S-shaped.", fig.height=3.4}
glm_age <- glm(treatment_uptake ~ age, data = dx, family = binomial)

# Predictions over a wide age range, on both scales
age_grid <- tibble(age = seq(15, 100, by = 1))
age_grid <- age_grid |>
  mutate(log_odds = predict(glm_age, newdata = age_grid, type = "link"),
         prob     = predict(glm_age, newdata = age_grid, type = "response"))

p_link <- ggplot(age_grid, aes(age, log_odds)) +
  geom_line(linewidth = 1, colour = "#0D7377") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(title = "Log-odds scale", x = "Age (years)",
       y = "Log-odds of uptake")

p_prob <- ggplot(age_grid, aes(age, prob)) +
  geom_line(linewidth = 1, colour = "#0D7377") +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(title = "Probability scale", x = "Age (years)",
       y = "Probability of uptake")

p_link + p_prob
```

### Estimation by maximum likelihood

Least squares is not appropriate for a binary outcome. Instead the
coefficients are estimated by *maximum likelihood*: we choose the values of
$\boldsymbol\beta$ that make the observed data most probable. For patient
$i$ the probability of the observed outcome is $p_i$ if $y_i = 1$ and
$1 - p_i$ if $y_i = 0$, which can be written compactly as
$p_i^{y_i}(1-p_i)^{1-y_i}$. Assuming independent patients, the likelihood
is the product over all patients, and it is easier to work with its
logarithm:

$$
\ell(\boldsymbol\beta) = \sum_{i=1}^{n}
\Big[\, y_i \log p_i + (1-y_i)\log(1-p_i) \Big],
$$

where each $p_i$ depends on $\boldsymbol\beta$ through the logistic
function. There is no closed-form solution; R maximises $\ell$ numerically
with an algorithm called *iteratively reweighted least squares* (also known
as Fisher scoring), which usually converges in a handful of iterations
[@agresti2013; @hosmer2013].

Two quantities derived from the likelihood appear in the R output and are
used for testing.

- The **deviance** of a model is $D = -2\ell(\hat{\boldsymbol\beta})$
  (relative to a saturated model that fits every observation perfectly).
  Smaller deviance means better fit. The *null deviance* is the deviance of
  a model with only an intercept; the *residual deviance* is that of the
  fitted model.
- The **likelihood-ratio (LR) test** compares two *nested* models, where
  the smaller model is a special case of the larger one. The difference in
  deviance, $D_{\text{small}} - D_{\text{large}}$, follows approximately a
  $\chi^2$ distribution with degrees of freedom equal to the number of extra
  parameters in the larger model.

Each individual coefficient is also accompanied by a **Wald test**,
$z = \hat\beta_j / \text{SE}(\hat\beta_j)$, compared with the standard normal
distribution. Wald and LR tests usually agree in large samples; when they
differ (small samples, sparse categories, very large effects) the LR test
is more reliable [@hosmer2013; @agresti2013].

### Interpreting the coefficients as odds ratios

Consider two patients who are identical in every predictor except $x_1$,
which is one unit higher for the second. Subtracting their linear
predictors, every term cancels except one:

$$
\operatorname{logit}(p_{\text{second}}) - \operatorname{logit}(p_{\text{first}})
= \beta_1 .
$$

A difference in log-odds is the log of an odds ratio, so

$$
\text{OR}_{x_1} = \exp(\beta_1).
$$

This is the central result for interpretation [@hosmer2013]:
**$\exp(\beta_j)$ is the odds ratio for a one-unit increase in $x_j$,
holding the other predictors fixed.** Because the model is linear on the
log-odds scale, this OR is the same whatever the starting value of $x_j$
and whatever the values of the other predictors (unless an interaction is
included; see Section 5.7).

How "one unit" is defined depends on the type of predictor.

- **Binary predictor** (for example diabetes, coded No = 0, Yes = 1).
  $\exp(\beta)$ is the OR comparing patients with diabetes to those without,
  the reference category.
- **Categorical predictor with several levels** (for example education:
  None, Primary, Secondary, Tertiary). R creates one indicator ("dummy")
  variable for each non-reference level. Each $\exp(\beta)$ compares that
  level with the reference level (None), not with the level below it.
- **Continuous predictor** (for example age in years). $\exp(\beta)$ is the
  OR per one-unit increase (per year). A one-year difference is rarely
  clinically interesting, so we usually report the OR for a meaningful
  increment $c$, such as 10 years:

$$
\text{OR}_{c\text{ units}} = \exp(c\,\beta) = \left[\exp(\beta)\right]^c ,
$$

with confidence interval $\exp\!\big(c\,\hat\beta \pm 1.96\, c\,
\text{SE}(\hat\beta)\big)$. Note that the OR for 10 years is the per-year OR
raised to the power 10, not multiplied by 10.

### Assumptions of logistic regression

Logistic regression makes fewer assumptions than linear regression, but
they are not trivial [@hosmer2013; @vittinghoff2012]:

1. **Correct link and linearity in the logit.** Each continuous predictor
   is linearly related to the *log-odds* of the outcome (not to the
   probability). This can be checked by adding non-linear terms.
2. **Independence.** Outcomes of different patients are independent given
   the predictors. Clustering of patients within facilities is a potential
   violation that we discuss in Section 5.8.
3. **No perfect separation.** No predictor (or combination) perfectly
   predicts the outcome. If, say, every patient with diabetes were on
   treatment, the maximum likelihood estimate of the diabetes coefficient
   would be infinite; R warns that "fitted probabilities numerically 0 or 1
   occurred" and reports a huge coefficient and standard error.
4. **Adequate sample size.** Maximum likelihood estimates are only
   approximately unbiased and normal in large samples; the number of
   *events* relative to the number of parameters matters (Section 5.8).

There is no assumption that the predictors are normally distributed, and
no assumption of equal variance.

## Fitting the logistic model in R {#sec-fit}

### A first model with one predictor

Logistic regression is fitted in R with `glm()`, the function for
generalised linear models, by specifying `family = binomial` (the logit
link is the default). The outcome can be a factor, in which case the first
level is treated as 0 (failure) and every other level as 1 (success).
We start with the crude model for health insurance and check that it
reproduces the odds ratio we computed by hand.

```{r c5-glm-crude}
glm_ins <- glm(treatment_uptake ~ health_insurance,
               data = dx, family = binomial(link = "logit"))
coef(glm_ins)          # log-odds scale
exp(coef(glm_ins))     # odds scale
```

The intercept, `r round(coef(glm_ins)[1], 3)`, is the log-odds of uptake in
the reference group (uninsured); exponentiated, it is the odds
`r round(exp(coef(glm_ins)[1]), 3)` that we saw in the table. The
coefficient for `health_insuranceYes`, `r round(coef(glm_ins)[2], 3)`, is
the difference in log-odds between insured and uninsured patients, and its
exponential, `r round(exp(coef(glm_ins)[2]), 2)`, is exactly the crude OR
computed by hand. A one-predictor logistic regression is simply a
re-expression of a $2 \times 2$ table.

### The multivariable model

Now we fit the main model of the study. Its predictors were chosen *before*
the analysis on clinical and public-health grounds (Section 5.8 explains
why this matters):

- demographic characteristics that commonly confound health-service use:
  age, sex, education and residence;
- clinical factors that increase contact with health services: diabetes
  and a family history of hypertension;
- the study's stated determinants: health insurance (financial access),
  hypertension knowledge score (awareness) and distance to the facility
  (geographical access).

```{r c5-glm-full}
model_full <- glm(
  treatment_uptake ~ age + sex + education + residence + diabetes +
    family_history_htn + health_insurance + knowledge_score +
    distance_to_facility_km,
  data   = dx,
  family = binomial(link = "logit")
)

summary(model_full)
```

### Reading the summary output line by line

This output contains a great deal of information, and it is worth reading
slowly.

- **Call.** The model formula, the family and the data set. Always check
  that the formula is what you intended.
- **Coefficients.** One row per estimated parameter, on the **log-odds
  scale**.
  - `Estimate` is $\hat\beta_j$. A positive value means higher odds of
    uptake as the predictor increases (or in that category compared with
    the reference); a negative value means lower odds.
  - `Std. Error` is the standard error of $\hat\beta_j$, a measure of its
    precision. Note how large it is for `diabetesYes` (0.474) compared with
    the other binary predictors (about 0.14–0.15). Only 28 diagnosed
    patients have diabetes, so this coefficient is estimated imprecisely.
  - `z value` is the Wald statistic, Estimate / Std. Error.
  - `Pr(>|z|)` is the two-sided Wald p-value for the null hypothesis that
    $\beta_j = 0$ (OR $= 1$), adjusted for the other variables.
- **Reading individual rows.**
  - `(Intercept)` $-3.77$ is the log-odds of uptake for a hypothetical
    patient with every predictor at zero or its reference level: aged 0,
    female, no education, rural, no diabetes, no family history, uninsured,
    knowledge score 0, living 0 km from the facility. It anchors the model
    but has no clinical meaning.
  - `age` 0.030: each additional year of age increases the log-odds of
    uptake by 0.030, other variables held fixed.
  - `sexMale` $-0.302$: men have lower log-odds of uptake than women of the
    same age, education, residence and so on. The suffix `Male` tells you
    that Female is the reference.
  - `educationPrimary`, `educationSecondary`, `educationTertiary`: each
    compares that level with no formal education. The coefficients
    increase steadily (0.33, 0.65, 0.89), suggesting a gradient.
  - `distance_to_facility_km` $-0.019$: lower odds with greater distance,
    but the p-value (0.128) indicates that the data are compatible with no
    association once the other variables are accounted for.
- **Signif. codes.** The stars are a legend for p-value thresholds. They
  encourage dichotomous thinking and should never be reported in a paper
  [@wasserstein2016; @greenland2016].
- **Dispersion parameter.** For a binomial model it is fixed at 1; this
  line matters only for other families.
- **Null deviance** (1369.1 on 991 degrees of freedom) is the deviance of an
  intercept-only model; **residual deviance** (1221.9 on 980 df) is the
  deviance after adding the 11 predictor terms. The drop of 147.2 on 11 df
  is the LR statistic for the null hypothesis that *none* of the predictors
  is associated with uptake, which is overwhelmingly rejected.
- **`(97 observations deleted due to missingness)`.** This line is easy to
  overlook and very important. `glm()` uses complete cases only: any patient
  with a missing value in *any* model variable is dropped. The model was
  fitted to 992 of the 1,089 diagnosed patients (degrees of freedom
  991 + 1). This number must appear in your paper.
- **AIC** (Akaike information criterion), $\text{AIC} = D + 2q$ where $q$ is
  the number of parameters, balances fit against complexity. It is useful
  for comparing models fitted to the *same* observations; on its own it has
  no interpretation.
- **Number of Fisher Scoring iterations** (4) shows that the algorithm
  converged quickly. A large number (say, more than 20) is a warning sign,
  often of separation.

```{r c5-nobs}
nobs(model_full)                          # observations used
table(model_full$y)                       # outcome among those used (0/1)
```

::: {.callout-warning title="Common mistake"}
Reporting the sample size of the data set rather than of the model. Here 97
diagnosed patients (8.9%) were excluded from the model because of missing
knowledge score, distance, education or age. Always report the number
analysed, describe the missing data, and consider whether a complete-case
analysis could be biased. When missingness is substantial, multiple
imputation is usually preferable [@sterne2009; @little2019].
:::

### Odds ratios with confidence intervals

The log-odds coefficients are needed for the mathematics, but readers need
odds ratios. `broom::tidy()` with `exponentiate = TRUE` and
`conf.int = TRUE` exponentiates the estimates and their confidence limits
in one step.

```{r c5-tidy-or}
or_table <- tidy(model_full, exponentiate = TRUE, conf.int = TRUE)

or_table |>
  select(term, OR = estimate, lower = conf.low, upper = conf.high,
         p = p.value) |>
  mutate(across(c(OR, lower, upper), \(x) round(x, 2)),
         p = format.pval(p, digits = 2, eps = 0.001))
```

The confidence limits returned by `tidy()` come from `confint()`, which for
a `glm` uses the *profile likelihood*. These intervals are slightly
asymmetric on the log scale and are more accurate than the Wald intervals
$\exp(\hat\beta \pm 1.96\,\text{SE})$, particularly for imprecise
estimates such as the diabetes OR [@hosmer2013]. Notice also that a
standard error cannot simply be exponentiated: the `std.error` column of
`tidy()` stays on the log-odds scale even when `exponentiate = TRUE`.

Each odds ratio is interpreted as an *adjusted* association, holding the
other variables in the model fixed.

- **Age**, OR 1.03 per year (95% CI 1.02 to 1.04): each additional year is
  associated with 3% higher odds of uptake.
- **Sex**, OR 0.74 (0.56 to 0.97): men have about 26% lower odds of uptake
  than women.
- **Education**: compared with no formal education, the ORs are 1.39
  (0.95 to 2.05) for primary, 1.92 (1.27 to 2.90) for secondary and 2.44
  (1.48 to 4.06) for tertiary education, a clear gradient.
- **Urban residence**, OR 1.87 (1.40 to 2.49) compared with rural.
- **Diabetes**, OR 3.56 (1.46 to 9.61): the largest OR, but also by far the
  widest interval, reflecting the small number of patients with diabetes.
- **Family history of hypertension**, OR 1.91 (1.44 to 2.53).
- **Health insurance**, OR 2.05 (1.54 to 2.74).
- **Knowledge score**, OR 1.10 per point (1.06 to 1.14).
- **Distance to facility**, OR 0.98 per km (0.96 to 1.01), p = 0.13.

::: {.callout-note title="Clinical interpretation"}
The width of a confidence interval tells you how much the data constrain
the estimate. The diabetes interval runs from a modest 1.5-fold to an
almost 10-fold increase in odds: we can be fairly confident the association
is positive, but we know little about its size. In contrast, the interval
for health insurance (1.5 to 2.7) is narrow enough to be useful for
planning. A large point estimate with a wide interval is weaker evidence
than a moderate estimate with a narrow one. (The data are simulated for
teaching; these results illustrate methods, not real clinical findings.)
:::

### Odds ratios for clinically meaningful increments

The per-unit ORs for age, knowledge and distance are close to 1 simply
because one year, one point or one kilometre is a small change. We rescale
them with $\exp(c\,\hat\beta)$, applying the same multiplication to the
confidence limits on the log scale.

```{r c5-rescale}
# Increments that are meaningful for each continuous predictor
increments <- tibble(
  term = c("age", "knowledge_score", "distance_to_facility_km"),
  unit = c("10 years", "5 points", "5 km"),
  c    = c(10, 5, 5)
)

# Profile-likelihood CIs on the log-odds scale, then rescale by c
ci_log <- confint(model_full)

rescaled <- increments |>
  mutate(beta  = coef(model_full)[term],
         OR    = exp(c * beta),
         lower = exp(c * ci_log[term, 1]),
         upper = exp(c * ci_log[term, 2])) |>
  select(term, unit, OR, lower, upper)

kable(rescaled, digits = 2,
      caption = "Adjusted odds ratios for clinically meaningful increments.")
```

Expressed this way the results become much easier to communicate: a
10-year difference in age is associated with
`r round(rescaled$OR[1], 2)` times the odds of uptake, a 5-point higher
knowledge score with `r round(rescaled$OR[2], 2)` times the odds, and living
5 km further away with `r round(rescaled$OR[3], 2)` times the odds (95% CI
`r round(rescaled$lower[3], 2)` to `r round(rescaled$upper[3], 2)`). The
increment should be chosen for clinical relevance (often the
interquartile range or a familiar unit such as a decade), stated clearly,
and decided before seeing the results.

### Predicted probabilities

Odds ratios are relative measures. To show what the model implies in
absolute terms, we can compute predicted probabilities for representative
patients with `predict(type = "response")`. Here we compare two
55-year-old women from a rural area with primary education, no diabetes,
no family history, an average knowledge score of 10 and living 6 km from
the facility, who differ only in health insurance.

```{r c5-predict}
profiles <- tibble(
  age = 55, sex = "Female", education = "Primary", residence = "Rural",
  diabetes = "No", family_history_htn = "No",
  health_insurance = c("No", "Yes"),
  knowledge_score = 10, distance_to_facility_km = 6
)

profiles |>
  mutate(p_uptake = predict(model_full, newdata = profiles,
                            type = "response")) |>
  select(health_insurance, p_uptake)
```

```{r c5-predict-values, include=FALSE}
pp <- predict(model_full, newdata = profiles, type = "response")
```

The predicted probability of uptake rises from `r round(100 * pp[1])`% to
`r round(100 * pp[2])`%, an absolute difference of
`r round(100 * (pp[2] - pp[1]))` percentage points. The odds ratio is the
same for every profile, but the absolute difference depends on the baseline
probability, which is why predicted probabilities (or risk differences) are
a valuable complement to ORs when communicating with clinicians and policy
makers.

### Testing a categorical predictor with several levels

Education enters the model as three indicator variables, each with its own
Wald p-value. These answer three separate questions (is Primary different
from None? Secondary from None? Tertiary from None?) but not the overall
question "is education associated with uptake?". For that we need a single
test of the null hypothesis that all three coefficients are zero, and the
likelihood-ratio test is the natural choice [@hosmer2013].

`drop1()` removes each term in turn (all levels of a factor together) and
compares the deviance with that of the full model.

```{r c5-drop1}
drop1(model_full, test = "LRT")
```

The row for education shows that removing it increases the deviance by
16.3 on 3 degrees of freedom, $p = 0.001$: education as a whole is
associated with uptake. The other rows give LR tests for each one-parameter
term; their p-values are very close to the Wald p-values in `summary()`,
as expected in a sample of this size.

The same test can be obtained by fitting the model without education and
comparing the two nested models directly with `anova()`. There is a trap
here. Education has 24 missing values among the diagnosed patients, so a
model *without* education, fitted to `dx`, would include patients who were
excluded from `model_full`. The two deviances would then refer to different
patients and could not be compared (R stops with the error "models were not
all fitted to the same size of dataset"). We therefore extract the
complete-case data used by the full model with `model.frame()` and fit the
smaller model to exactly those patients.

```{r c5-anova-lrt}
cc <- model.frame(model_full)       # the 992 patients used by model_full
nrow(cc)

model_no_edu <- update(model_full, . ~ . - education, data = cc)
lrt_edu <- anova(model_no_edu, model_full, test = "LRT")
attr(lrt_edu, "heading") <- NULL   # omit the long model formulas
lrt_edu
```

The LR statistic (the difference in deviance, 16.3 on 3 df) and its
p-value are identical to the `drop1()` result, which performs the same
comparison internally.

::: {.callout-tip title="Good practice"}
When you compare nested models, make sure both were fitted to exactly the
same observations. The simplest way is to create the complete-case data set
once, with `model.frame()` of the largest model or with `drop_na()` on all
the candidate variables, and to fit every model to it.
:::

## Confounding: crude versus adjusted estimates {#sec-confounding}

### What is confounding?

A *confounder* is a variable that distorts the observed association
between an exposure and an outcome because it is associated with both.
The classical criteria are that a confounder (1) is a cause, or a proxy
for a cause, of the outcome; (2) is associated with the exposure in the
source population; and (3) is not an effect of the exposure, that is, it
does not lie on the causal pathway from exposure to outcome
[@rothman2008]. The *crude* (unadjusted) estimate mixes the effect of the
exposure with that of the confounder; the *adjusted* estimate, from a model
that includes the confounder, attempts to separate them.

### Causal diagrams in words

Causal diagrams, or directed acyclic graphs (DAGs), make these ideas
precise [@greenland1999dag]. A DAG draws each variable as a node and each
assumed direct causal effect as an arrow. Consider the question "is
distance to the facility associated with treatment uptake?". Plausible
assumptions are:

- *Distance → Uptake*: patients who live far from a clinic face travel
  costs and time, which may reduce uptake.
- *Residence → Distance*: rural patients tend to live further from the
  facility.
- *Residence → Uptake*: independently of distance, urban areas may have
  more pharmacies, more health information and different patterns of care.

Residence therefore opens a "back-door path" from distance to uptake
(Distance ← Residence → Uptake) that is not part of the effect of distance.
To estimate the association of distance with uptake that is not due to
residence, the back-door path must be blocked by adjusting for residence.
In contrast, a variable caused by the exposure and lying on the path to the
outcome (a *mediator*) should not be adjusted for if we want the total
effect, and adjusting for a common *effect* of exposure and outcome (a
*collider*) can create bias where none existed [@greenland1999dag;
@vanderweele2019].

@vanderweele2019 offers a practical rule for situations in which the full
causal structure is uncertain, the *modified disjunctive cause criterion*:
adjust for any pre-exposure variable that is a cause of the exposure, the
outcome or both; exclude known instrumental variables (causes of the
exposure only); and include good proxies for unmeasured common causes.

### A worked example: distance and residence

The data agree with the first two arrows: rural patients in the model
sample live on average 9.6 km from their facility, urban patients 6.0 km.
We now compare the crude odds ratio for distance with the odds ratio
adjusted for residence alone and for all the variables in the full model.
To make the comparison fair, all three models are fitted to the same 992
complete cases.

```{r c5-confounding}
# cc: the complete-case data used by the full model (created above)
with(cc, tapply(distance_to_facility_km, residence, mean))

m_crude <- glm(treatment_uptake ~ distance_to_facility_km,
               data = cc, family = binomial)
m_res   <- glm(treatment_uptake ~ distance_to_facility_km + residence,
               data = cc, family = binomial)

# Extract the OR for distance (per 5 km) and its 95% CI from each model
dist_or <- function(model, label) {
  b  <- coef(model)["distance_to_facility_km"]
  ci <- confint(model)["distance_to_facility_km", ]
  p  <- summary(model)$coefficients["distance_to_facility_km", 4]
  tibble(model = label, OR_5km = exp(5 * b),
         lower = exp(5 * ci[1]), upper = exp(5 * ci[2]), p = p)
}

conf_tab <- bind_rows(
  dist_or(m_crude,    "Crude"),
  dist_or(m_res,      "Adjusted for residence"),
  dist_or(model_full, "Fully adjusted")
)
kable(conf_tab, digits = 3,
      caption = "OR for uptake per 5 km of distance: crude and adjusted.")
```

```{r c5-confounding-values, include=FALSE}
chg <- 100 * (log(conf_tab$OR_5km[2]) - log(conf_tab$OR_5km[1])) /
  abs(log(conf_tab$OR_5km[1]))
```

The crude analysis suggests that each additional 5 km is associated with
`r round(100 * (1 - conf_tab$OR_5km[1]))`% lower odds of uptake, with a
confidence interval that excludes 1 ($p = `r signif(conf_tab$p[1], 2)`$).
After adjusting for residence alone, the OR moves towards 1
(`r round(conf_tab$OR_5km[2], 2)`), the interval includes 1, and the
p-value rises to `r round(conf_tab$p[2], 2)`. Expressed on the log-odds
scale, the coefficient has shrunk by about `r round(abs(chg))`%. Much of the
crude association between distance and uptake was due to residence: rural
patients live further away *and*, for reasons other than distance, are less
likely to be on treatment. Adjusting for the remaining variables changes
the distance OR very little further.

A common rule of thumb flags a variable as a confounder if adjusting for it
changes the coefficient of the exposure by more than 10% (the
*change-in-estimate* criterion). It is a useful screening tool, but it
should not replace causal reasoning: a mediator or a collider can also
change an estimate, and adjusting for it would introduce bias
[@rothman2008; @vanderweele2019].

::: {.callout-note title="Clinical interpretation"}
The crude analysis might have led a health planner to conclude that distance
itself is a major barrier to treatment. The adjusted analysis suggests a
more nuanced picture: rural residence is associated with lower uptake, and
distance carries little additional information once residence is known.
Interventions targeting rural patients (for example outreach or community
dispensing) might therefore matter more than simply reducing travel
distance. In the real world this would need confirmation from better
designed studies; these data are simulated.
:::

### Not every change is confounding: non-collapsibility

Compare the crude and adjusted ORs for two other variables in the same 992
patients.

```{r c5-noncollapse}
crude_or <- function(var) {
  f <- reformulate(var, response = "treatment_uptake")
  exp(coef(glm(f, data = cc, family = binomial)))[2]
}
tibble(variable = c("family_history_htn", "health_insurance"),
       crude_OR = c(crude_or("family_history_htn"),
                    crude_or("health_insurance")),
       adjusted_OR = exp(coef(model_full)[c("family_history_htnYes",
                                            "health_insuranceYes")])) |>
  kable(digits = 2, caption = "Crude and adjusted ORs for two predictors.")
```

Both odds ratios move *away* from 1 after adjustment, from about 1.65 to
about 2.0, although family history and insurance are only weakly related
to the other predictors in these data. Only part of this change is
confounding. The rest is a mathematical property of the odds ratio called
*non-collapsibility*: when other variables that strongly predict the
outcome are added to a logistic model, the OR for a given exposure
generally moves away from the null *even in the complete absence of
confounding* [@rothman2008; @vittinghoff2012]. The adjusted
(*conditional*) OR compares patients with identical values of all the
other predictors, whereas the crude (*marginal*) OR averages over the
population, and with odds ratios these two quantities differ. Risk
differences and risk ratios do not have this property. Exercise 5.4 shows
how to separate the two components by standardisation. The practical
lesson is that a change between crude and adjusted ORs is not, by itself,
proof of confounding; your causal reasoning must come first.

### The Table 2 fallacy

It is tempting to interpret *every* adjusted OR in a multivariable model as
the effect of that variable, as we did, with caveats, in the list above.
@westreich2013 called this the *Table 2 fallacy* (since the main
multivariable results are often in a paper's second table). The problem is
that a model built to estimate the effect of one exposure is adjusted for
that exposure's confounders, which are generally not the right set of
adjustment variables for each of the other predictors.

Take education and knowledge score. Education plausibly *causes* better
knowledge of hypertension, which in turn increases uptake. In our model,
which includes knowledge, the education OR estimates the association of
education with uptake *among patients with the same knowledge score*: the
part of the education effect that does not operate through knowledge (the
*direct effect*). The total effect of education would be estimated by a
model that omits its mediators. Similarly, the OR for age is adjusted for
diabetes, which is itself partly a consequence of age.

::: {.callout-warning title="Common mistake"}
Interpreting each coefficient in a "determinants" model as a causal effect
of equal standing. In a descriptive multivariable model, the ORs are
*mutually adjusted associations*. If one particular exposure is of causal
interest, build the model for that exposure (using a causal diagram) and
report only its OR as the main result [@westreich2013].
:::

### Interaction (effect modification)

So far we have assumed that the OR for each predictor is the same at every
value of the other predictors. *Interaction*, or *effect modification*,
occurs when the association of one variable with the outcome differs
according to the level of another [@rothman2008; @kirkwood2003]. For
example, the association of diabetes with uptake might be stronger in
younger patients, for whom a diabetes diagnosis brings regular clinic
contact. An interaction is modelled by adding the product of the two
variables:

$$
\operatorname{logit}(p) = \beta_0 + \beta_1\,\text{age} +
\beta_2\,\text{diabetes} + \beta_3\,(\text{age} \times \text{diabetes}) +
\dots
$$

The OR for diabetes is then $\exp(\beta_2 + \beta_3 \times \text{age})$,
which changes with age. We test $\beta_3 = 0$ with an LR test comparing
the models with and without the product term.

```{r c5-interaction}
model_int <- update(model_full, . ~ . + diabetes:age)
anova(model_full, model_int, test = "LRT")
```

The interaction term reduces the deviance by only 1.3 on 1 df ($p = 0.25$).
There is no evidence that the diabetes OR varies with age, so we keep the
simpler model without the interaction. Interactions should be specified in
advance on clinical grounds and tested sparingly: tests of interaction have
low power, and searching many possible interactions produces spurious
findings.

::: {.callout-warning title="Common mistake"}
When an interaction term *is* kept in a model, the "main effect"
coefficients change meaning. In the model above, `diabetesYes` would be the
log OR for diabetes at age 0, a meaningless value. Report the OR for one
variable at chosen values of the other (for example at ages 40, 55 and
70), or centre the continuous variable before forming the product.
:::

## Building the model {#sec-model-building}

### Pre-specification versus automated selection

How should the predictors in a multivariable model be chosen? In an
explanatory or descriptive analysis the answer is: from subject-matter
knowledge, decided before looking at the association of each candidate
with the outcome, and documented in a statistical analysis plan
[@vittinghoff2012; @harrell2015; @heinze2018]. This is how
`model_full` was built.

The alternative, still common in clinical papers, is to let an algorithm
choose: *forward selection* adds variables one at a time, *backward
elimination* starts with all candidates and removes the least significant,
and *stepwise* methods alternate between the two. A variant is the popular
two-step approach of screening variables with univariable tests (for
example keeping those with $p < 0.2$) and then running backward
elimination. These procedures have well-documented problems
[@harrell2015; @heinze2018]:

- **Biased estimates.** Variables are selected *because* their estimated
  effects happen to be large in this sample, so the selected coefficients
  are biased away from zero.
- **Invalid inference.** Standard errors, confidence intervals and
  p-values of the final model ignore the selection process; intervals are
  too narrow and p-values too small.
- **Instability.** Small changes to the data (a different sample, a few
  extra patients) often produce a different set of "selected" variables.
- **Dropped confounders.** An important confounder may be removed because
  its own p-value is above 0.05, reintroducing the bias that adjustment was
  meant to remove.
- **Univariable screening is misleading.** A variable can be unassociated
  with the outcome on its own but important after adjustment (and vice
  versa), so crude p-values are a poor guide to what belongs in a
  multivariable model.

R's `step()` function performs stepwise selection by AIC. For completeness,
here is what it does with our model.

```{r c5-step}
model_step <- step(model_full, direction = "both", trace = 0)
attr(terms(model_step), "term.labels")      # the variables it kept
AIC(model_full, model_step)
```

In this data set `step()` keeps every variable, including distance with its
p-value of 0.13, because removing distance would slightly increase the AIC.
On a different data set it might drop several variables. Our decision does
not depend on what `step()` says: the pre-specified model is the main
analysis.

::: {.callout-tip title="Good practice"}
If you do use data-driven selection (for example in a prediction model),
use methods designed for it, such as penalised regression, report the
procedure fully, and assess the stability of the selection with
bootstrap resampling [@heinze2018; @harrell2015]. Never present a model
chosen by stepwise selection as if it had been specified in advance.
:::

### How many predictors can the data support?

Every parameter in a logistic model "uses up" information. The relevant
amount of information is not the total sample size but the number of
patients in the *less frequent* outcome category, often called the number
of events. @peduzzi1996 showed in simulations that with fewer than about
10 *events per variable* (EPV, strictly per estimated parameter) the
coefficients of a logistic model can be biased, their variances poorly
estimated, and confidence intervals unreliable. The "10 EPV" rule is a
rough guide rather than a law, and prediction models in particular may
need considerably more [@harrell2015], but it is a useful sanity check.

```{r c5-epv}
n_events     <- min(table(model_full$y))     # less frequent category
n_parameters <- length(coef(model_full)) - 1 # excluding the intercept
c(events = n_events, parameters = n_parameters,
  EPV = round(n_events / n_parameters, 1))
```

With `r n_events` events (patients on treatment) and `r n_parameters`
parameters, the EPV is about `r round(n_events / n_parameters)`, well above
10. The model is not overfitted. The imprecision of the diabetes estimate
comes from a different problem: there are only 28 diagnosed patients with
diabetes, and the precision of a binary predictor's OR depends on the
number of patients in its smallest cell, not on the EPV of the whole
model.

### Collinearity

*Collinearity* (or multicollinearity) occurs when predictors are strongly
correlated with each other. The model then struggles to separate their
individual contributions, standard errors become inflated, and estimates
can change wildly when a variable is added or removed. The *variance
inflation factor* (VIF) for predictor $j$ is

$$
\text{VIF}_j = \frac{1}{1 - R_j^2},
$$

where $R_j^2$ is the $R^2$ from regressing $x_j$ on all the other
predictors. A VIF of 1 means no collinearity; values above 5, and
especially above 10, indicate a problem [@vittinghoff2012]. For factors
with several levels `car::vif()` reports the *generalised* VIF (GVIF) and
$\text{GVIF}^{1/(2\,\text{df})}$, whose square is comparable with the usual
VIF thresholds.

```{r c5-vif}
car::vif(model_full)
```

All values are close to 1 (the largest GVIF, for residence, is about 1.12).
The predictors carry largely separate information, and collinearity is not
a concern. In clinical data, collinearity typically arises from including
two measures of the same thing (weight and BMI, systolic and mean arterial
pressure) or a variable together with a transformation of it.

### Missing data and clustering

Two further design issues affect this model. First, as noted, 97 patients
were excluded because of missing data. A complete-case analysis is unbiased
if missingness does not depend on the outcome given the predictors, but can
be biased otherwise; with about 9% of patients excluded it is reasonable to
report the complete-case analysis as primary and, in a full study, to
check it with multiple imputation [@sterne2009; @little2019].

Second, patients were recruited from six facilities. Patients attending the
same facility share staff, drug supplies and local policies, so their
outcomes may be correlated, violating the independence assumption. Ignoring
such clustering usually makes standard errors too small. Options include
adding facility as a fixed effect (feasible with only six facilities; see
Exercise 5.6), using cluster-robust standard errors, or fitting a mixed
model with a random facility effect; the last two are beyond the scope of
this book [@vittinghoff2012].

## Checking the model {#sec-checking}

A fitted model should be checked before its results are reported. For
logistic regression the main checks concern the form of the continuous
predictors, how well the model separates patients with and without the
outcome (*discrimination*), how closely predicted probabilities match
observed proportions (*calibration*), and whether a few observations unduly
influence the results [@hosmer2013; @harrell2015].

### Linearity of continuous predictors in the logit

The model assumes that age, knowledge score and distance each have a
linear relationship with the log-odds of uptake. A simple check is to add a
quadratic term for each and test it with an LR test. If the quadratic term
is unnecessary, the linear form is adequate.

```{r c5-linearity}
model_quad <- update(model_full, . ~ . + I(age^2) + I(knowledge_score^2) +
                       I(distance_to_facility_km^2))
anova(model_full, model_quad, test = "LRT")
```

The three quadratic terms together reduce the deviance by only a small
amount ($p = `r round(anova(model_full, model_quad, test = "LRT")[2, 5], 2)`$),
so there is no evidence of curvature and the linear terms are retained.
More flexible checks use restricted cubic splines [@harrell2015].

### Discrimination: the ROC curve and the AUC

For each patient the model produces a predicted probability of uptake.
If we classify a patient as a "predicted taker" whenever this probability
exceeds a threshold $t$, we can compute the *sensitivity* (the proportion
of actual takers classified correctly) and the *specificity* (the
proportion of non-takers classified correctly). The *receiver operating
characteristic* (ROC) curve plots sensitivity against $1 -$ specificity for
every possible threshold.

The *area under the ROC curve* (AUC, also called the $c$ statistic)
summarises discrimination in a single number. @hanley1982 showed that it
has a simple interpretation: the AUC is the probability that a randomly
chosen patient *with* the outcome has a higher predicted probability than a
randomly chosen patient *without* it. An AUC of 0.5 means the model
discriminates no better than tossing a coin, and 1.0 means perfect
separation. Common descriptive labels are 0.7–0.8 "acceptable", 0.8–0.9
"excellent" and above 0.9 "outstanding" [@hosmer2013].

```{r c5-roc}
roc_obj <- roc(response  = model_full$y,       # observed 0/1 outcome
               predictor = fitted(model_full),  # predicted probabilities
               quiet     = TRUE)
auc(roc_obj)
ci.auc(roc_obj)           # DeLong 95% confidence interval
```

```{r c5-roc-plot, fig.cap="ROC curve for the multivariable logistic regression model of treatment uptake. The dashed diagonal shows a model with no discrimination (AUC = 0.5).", fig.width=5, fig.height=4.5}
roc_df <- tibble(fpr = 1 - roc_obj$specificities,
                 tpr = roc_obj$sensitivities) |>
  arrange(fpr, tpr)

ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              colour = "grey50") +
  geom_step(linewidth = 0.9, colour = "#0D7377") +
  annotate("text", x = 0.65, y = 0.25,
           label = paste0("AUC = ", round(auc(roc_obj), 3))) +
  coord_equal() +
  labs(title = "ROC curve: treatment uptake model",
       x = "1 - specificity (false positive rate)",
       y = "Sensitivity (true positive rate)")
```

```{r c5-auc-values, include=FALSE}
auc_val <- as.numeric(auc(roc_obj))
auc_ci  <- as.numeric(ci.auc(roc_obj))
```

The AUC is `r round(auc_val, 2)` (95% CI `r round(auc_ci[1], 2)` to
`r round(auc_ci[3], 2)`): if we pick at random one patient on treatment and
one not on treatment, the model assigns the higher probability to the
treated patient about `r round(100 * auc_val)`% of the time. This is
acceptable discrimination for a model whose purpose is explanation rather
than individual prediction.

::: {.callout-note title="Clinical interpretation"}
An AUC of about 0.71 means that the measured determinants explain part, but
far from all, of why some patients take up treatment and others do not. The
model is useful for understanding the population; it would not be good
enough to decide which individual patients need extra support. Note also
that the AUC is computed on the same data used to fit the model, so it is
slightly optimistic; a prediction model would need internal validation, for
example by bootstrapping [@harrell2015].
:::

### Calibration and the Hosmer–Lemeshow test

Discrimination is about ranking; *calibration* is about agreement. A
well-calibrated model that predicts a 30% probability of uptake for a group
of patients should see about 30% of them on treatment. The classical check
groups patients by their predicted probability, usually into ten groups
("deciles of risk"), and compares the observed number of events $O_g$ in
each group with the expected number $E_g = \sum_{i \in g} \hat p_i$.

@hosmer1980 combined these comparisons into a goodness-of-fit statistic,

$$
\hat C = \sum_{g=1}^{G} \frac{(O_g - E_g)^2}{E_g\,(1 - E_g / n_g)},
$$

where $n_g$ is the number of patients in group $g$. If the model fits, $\hat
C$ approximately follows a $\chi^2$ distribution with $G - 2$ degrees of
freedom. A *small* p-value indicates poor fit. The test is available in
add-on packages, but it is simple to compute directly, which also makes
clear what it does.

```{r c5-hosmer-lemeshow}
calib <- tibble(y = model_full$y, p_hat = fitted(model_full)) |>
  mutate(decile = ntile(p_hat, 10)) |>          # 10 groups of equal size
  group_by(decile) |>
  summarise(n         = n(),
            mean_pred = mean(p_hat),            # mean predicted probability
            obs_prop  = mean(y),                # observed proportion
            expected  = sum(p_hat),             # E_g
            observed  = sum(y))                 # O_g

kable(calib, digits = 3,
      caption = "Observed and expected uptake by decile of predicted risk.")

hl_stat <- with(calib, sum((observed - expected)^2 /
                             (expected * (1 - expected / n))))
hl_df   <- nrow(calib) - 2
hl_p    <- pchisq(hl_stat, df = hl_df, lower.tail = FALSE)
c(statistic = round(hl_stat, 2), df = hl_df, p = round(hl_p, 3))
```

Across the ten groups, the observed and expected counts agree closely, and
the Hosmer–Lemeshow statistic is `r round(hl_stat, 2)` on `r hl_df` df
($p = `r round(hl_p, 2)`$): no evidence of poor calibration. The same
information is easier to see in a calibration plot.

```{r c5-calibration-plot, fig.cap="Calibration plot: observed proportion on treatment against mean predicted probability in ten groups of predicted risk. Points on the dashed line indicate perfect calibration.", fig.width=5, fig.height=4.5}
ggplot(calib, aes(x = mean_pred, y = obs_prop)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              colour = "grey50") +
  geom_line(colour = "#0D7377") +
  geom_point(size = 2.5, colour = "#0D7377") +
  scale_x_continuous(labels = percent, limits = c(0, 1)) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  coord_equal() +
  labs(title = "Calibration by decile of predicted risk",
       x = "Mean predicted probability of uptake",
       y = "Observed proportion on treatment")
```

::: {.callout-warning title="Common mistake"}
Treating a non-significant Hosmer–Lemeshow test as proof of a good model.
The test has low power in small samples and, in very large samples, flags
trivial miscalibration; its result also depends on the number of groups
chosen [@hosmer2013]. Use it alongside a calibration plot, and remember
that calibration in the data used to build the model says little about
calibration in a new population.
:::

### Influential observations

A single unusual patient, or a few, can noticeably change a regression
estimate. *Cook's distance* measures how much all the fitted values change
when observation $i$ is removed; `broom::augment()` computes it for every
observation. A common screening threshold is $4/n$.

```{r c5-cooks}
aug_glm <- augment(model_full)                # one row per patient used
cut_off <- 4 / nrow(aug_glm)
sum(aug_glm$.cooksd > cut_off)                # number above the threshold
max(aug_glm$.cooksd)                          # the largest Cook's distance

# Who are the 10 most influential patients?
top10 <- order(aug_glm$.cooksd, decreasing = TRUE)[1:10]
count(cc[top10, ], diabetes, treatment_uptake)
```

`r sum(aug_glm$.cooksd > cut_off)` of the 992 patients exceed the
threshold, but the largest Cook's distance is only
`r round(max(aug_glm$.cooksd), 3)`, far below the value of 1 that is
sometimes used to indicate a truly dominating observation. Who are the most
influential patients? Most of the ten largest values belong to patients
with diabetes who are *not* on treatment. This makes sense: only 7 of the
28 diagnosed patients with diabetes are untreated, so each of them carries
a great deal of the information about the diabetes coefficient.

What happens if we remove these ten patients, as an over-zealous analyst
might do to "clean" the model?

```{r c5-influence-sensitivity}
model_sens <- update(model_full, data = cc[-top10, ])
tibble(term = names(coef(model_full))[-1],
       full = exp(coef(model_full))[-1],
       without_top10 = exp(coef(model_sens))[-1]) |>
  mutate(full = round(full, 2), without_top10 = signif(without_top10, 3))
```

Most ORs barely change, but the diabetes OR explodes to an absurd value of
about 15 million. Once the untreated patients with diabetes in the data are
removed, *every* remaining patient with diabetes is on treatment: diabetes
then predicts the outcome perfectly. This is *complete separation*
(Section 5.5), and the maximum likelihood estimate of the diabetes
coefficient no longer exists. The example teaches two lessons. First,
influential observations are often genuine and informative patients, not
errors; deleting them can destroy the very information a coefficient
depends on. Second, flagged observations should be *inspected* (are they
data-entry errors, or real but unusual patients?), and any sensitivity
analysis should be reported alongside the main analysis rather than
replacing it.

## Presenting the results {#sec-presenting}

### A publication-ready regression table

The standard way to present a multivariable logistic regression in a paper
is a table with, for each predictor, the crude (univariable) OR and the
adjusted (multivariable) OR, each with its 95% CI and p-value. Showing
both lets readers see how much adjustment changes each estimate. Copying
numbers from the console into a word processor is slow and error-prone;
`gtsummary` builds the table directly from the model objects
[@sjoberg2021].

`tbl_uvregression()` fits one univariable model per predictor,
`tbl_regression()` formats the multivariable model, and `tbl_merge()` puts
them side by side. We fit the univariable models to the same 992 complete
cases (`cc`) so that crude and adjusted estimates are directly comparable.

```{r c5-gtsummary-table}
# Readable labels for the predictors
labels_list <- list(
  age ~ "Age (years)",
  sex ~ "Sex",
  education ~ "Education",
  residence ~ "Residence",
  diabetes ~ "Diabetes",
  family_history_htn ~ "Family history of hypertension",
  health_insurance ~ "Health insurance",
  knowledge_score ~ "Knowledge score (per point)",
  distance_to_facility_km ~ "Distance to facility (per km)"
)

# Univariable (crude) ORs, one model per predictor
tbl_uni <- tbl_uvregression(
  data         = cc,
  method       = glm,
  y            = treatment_uptake,
  method.args  = list(family = binomial),
  exponentiate = TRUE,
  label        = labels_list,
  hide_n       = TRUE
) |>
  modify_header(estimate ~ "**cOR**")

# Multivariable (adjusted) ORs from the pre-specified model
tbl_multi <- tbl_regression(model_full, exponentiate = TRUE,
                            label = labels_list) |>
  modify_header(estimate ~ "**aOR**")

tbl_both <- tbl_merge(list(tbl_uni, tbl_multi),
                      tab_spanner = c("**Crude**", "**Adjusted**"))

tbl_both |>
  as_kable(caption = paste("Crude and adjusted odds ratios for",
                           "hypertension treatment uptake (n = 992)."))
```

The table has seven columns: the characteristic, then OR, 95% CI and
p-value for the crude and adjusted analyses. Reference categories are shown
with a dash, which reminds the reader what each OR is compared with. In a
manuscript you would add a footnote: "cOR = crude odds ratio; aOR = odds
ratio adjusted for all other variables in the table; CI = confidence
interval. Complete-case analysis of 992 of 1,089 patients with diagnosed
hypertension."

::: {.callout-tip title="Good practice"}
In a manuscript, export the same table to Word or HTML with
`gtsummary::as_flex_table()` or `as_gt()`, so that the numbers in the paper
come directly from the code. If the data change, re-run the script and the
table updates itself. Report p-values to two significant figures (and
"<0.001" for very small values), never as stars or as "NS".
:::

### A forest plot of adjusted odds ratios

A *forest plot* displays each OR as a point with a horizontal line for its
confidence interval, and a vertical reference line at OR = 1. Because ORs
are ratios, the horizontal axis should be on a **logarithmic** scale: on a
linear scale an OR of 0.5 (halving the odds) would appear much closer to 1
than an OR of 2 (doubling), although the two effects are of the same
magnitude, and confidence intervals would look artificially asymmetric.

```{r c5-forest-plot, fig.cap="Forest plot of adjusted odds ratios (95% confidence intervals) for treatment uptake from the multivariable logistic regression model (n = 992). The horizontal axis is on a logarithmic scale; the dashed line marks OR = 1.", fig.height=4.8}
# Readable labels for each model term
term_labels <- c(
  age                     = "Age (per year)",
  sexMale                 = "Male (vs female)",
  educationPrimary        = "Primary (vs none)",
  educationSecondary      = "Secondary (vs none)",
  educationTertiary       = "Tertiary (vs none)",
  residenceUrban          = "Urban (vs rural)",
  diabetesYes             = "Diabetes",
  family_history_htnYes   = "Family history of HTN",
  health_insuranceYes     = "Health insurance",
  knowledge_score         = "Knowledge score (per point)",
  distance_to_facility_km = "Distance (per km)"
)

forest_df <- or_table |>
  filter(term != "(Intercept)") |>
  mutate(label = factor(term_labels[term],
                        levels = rev(term_labels)))  # keep model order

ggplot(forest_df, aes(x = estimate, y = label)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high),
                width = 0.25, colour = "#0D7377", orientation = "y") +
  geom_point(size = 2.6, colour = "#0D7377") +
  scale_x_log10(breaks = c(0.5, 0.75, 1, 1.5, 2, 3, 5, 10)) +
  labs(title = "Adjusted odds ratios for treatment uptake",
       x = "Adjusted odds ratio (log scale)", y = NULL)
```

The plot conveys the main messages at a glance: all the ORs except those
for sex and distance lie to the right of 1; most intervals exclude 1; the
diabetes interval is much the widest; and the distance OR sits just to the
left of 1 with an interval that crosses it. Note that the per-unit ORs for
continuous variables look small only because the units are small; a
forest plot that mixes per-unit and per-category ORs should say so in its
labels, or use rescaled ORs.

## Writing the results {#sec-writing}

### What to report

The STROBE statement (*Strengthening the Reporting of Observational
Studies in Epidemiology*) lists the items that reports of cohort,
case–control and cross-sectional studies should contain [@vonelm2007].
For a regression analysis, the most relevant are:

- **Participants**: the number of individuals at each stage, including the
  number eligible, the number analysed, and the reasons for exclusion (here
  1,089 diagnosed patients, of whom 992 had complete data).
- **Statistical methods**: the model, how variables were chosen, how
  continuous variables were handled, how missing data were addressed, and
  any sensitivity analyses.
- **Main results**: crude and adjusted estimates with their precision (95%
  CIs), making clear which confounders were adjusted for and why.
- **Category boundaries and units**: for continuous predictors, the unit
  to which each OR refers.
- **Other analyses**: interactions examined and sensitivity analyses.

### A Methods paragraph

> We used multivariable logistic regression to estimate the association
> of patient and access characteristics with treatment uptake among adults
> with diagnosed hypertension. Predictors were specified before analysis on
> the basis of the literature and clinical reasoning: age, sex, education,
> residence, diabetes, family history of hypertension, health insurance,
> hypertension knowledge score and distance to the facility. Continuous
> predictors were modelled as linear terms after checking for non-linearity
> with quadratic terms. Results are reported as odds ratios (ORs) with 95%
> profile-likelihood confidence intervals; overall associations of
> categorical variables were assessed with likelihood-ratio tests.
> Discrimination was assessed with the area under the ROC curve and
> calibration with the Hosmer–Lemeshow test. Analyses used complete cases
> and were performed in R version 4.6.

```{r c5-results-values, include=FALSE}
n_dx      <- nrow(dx)
n_uptake  <- sum(dx$treatment_uptake == "Yes")
n_model   <- nobs(model_full)
fmt_or <- function(term) {
  r <- or_table[or_table$term == term, ]
  sprintf("%.2f, 95%% CI %.2f to %.2f", r$estimate, r$conf.low, r$conf.high)
}
p_edu <- drop1(model_full, test = "LRT")["education", "Pr(>Chi)"]
```

### A Results paragraph

The paragraph below is written with inline R code, so every number comes
straight from the model objects created above.

> Of the `r format(n_dx, big.mark = ",")` adults with diagnosed
> hypertension, `r n_uptake` (`r round(100 * n_uptake / n_dx, 1)`%) were
> receiving antihypertensive treatment; `r n_model` had complete data and
> were included in the multivariable model. After mutual adjustment,
> treatment uptake was associated with diabetes (OR
> `r fmt_or("diabetesYes")`), health insurance (OR
> `r fmt_or("health_insuranceYes")`), a family history of hypertension (OR
> `r fmt_or("family_history_htnYes")`) and urban rather than rural
> residence (OR `r fmt_or("residenceUrban")`). The odds of uptake increased
> with educational attainment (likelihood-ratio $p =
> `r round(p_edu, 3)`$; OR for tertiary versus no education
> `r fmt_or("educationTertiary")`), with age (OR per 10 years
> `r sprintf("%.2f, 95%% CI %.2f to %.2f", rescaled$OR[1], rescaled$lower[1], rescaled$upper[1])`)
> and with hypertension knowledge (OR per 5 points
> `r sprintf("%.2f, 95%% CI %.2f to %.2f", rescaled$OR[2], rescaled$lower[2], rescaled$upper[2])`).
> Men had lower odds of uptake than women (OR `r fmt_or("sexMale")`).
> Distance to the facility was associated with lower uptake in the crude
> analysis, but after adjustment (mainly for residence) the association was
> weaker and compatible with no association (OR per 5 km
> `r sprintf("%.2f, 95%% CI %.2f to %.2f", rescaled$OR[3], rescaled$lower[3], rescaled$upper[3])`).
> The model showed acceptable discrimination (AUC
> `r sprintf("%.2f, 95%% CI %.2f to %.2f", auc_val, auc_ci[1], auc_ci[3])`)
> and no evidence of poor calibration (Hosmer–Lemeshow $p =
> `r round(hl_p, 2)`$).

Notice the features of this paragraph. It starts with the participants and
the number analysed. Every OR has a 95% CI, and the reference category is
clear ("urban rather than rural", "men ... than women"). Continuous
predictors are reported per meaningful increment. A multi-level variable
is summarised by its overall LR test plus the most informative comparison.
A non-significant result (distance) is reported with its estimate and
interval, not as "not significant" alone. And the wording describes
*associations*, not effects.

::: {.callout-warning title="Common mistake"}
Over-claiming causality. This is a cross-sectional study: exposure and
outcome were measured at the same time, so we cannot tell whether, for
example, knowledge led to treatment or treatment (and the counselling that
accompanies it) led to knowledge. Words such as "effect", "impact",
"leads to", "reduces" and "determines" imply causation. Prefer "was
associated with", "had higher odds of" and "was more common among", and
discuss reverse causation and residual confounding in the Discussion
[@vonelm2007].
:::

::: {.callout-note title="Clinical interpretation"}
If these findings were real, they would point to several groups at risk of
not receiving treatment after diagnosis: men, rural residents, people
without health insurance, those with little formal education and those with
poor knowledge of hypertension. Some of these factors (insurance coverage,
knowledge) are potentially modifiable through policy and health education.
Remember, however, that the data in this book are simulated for teaching:
the results illustrate methods, not real clinical findings.
:::

### Results, interpretation and discussion

A Results section reports what was found: numbers, intervals, and neutral
wording. The interpretation of those numbers (what they mean for patients
and services), the comparison with previous studies, and the limitations
belong in the Discussion. For this analysis the limitations section would
mention the cross-sectional design (no temporal ordering), self-reported
uptake, the complete-case analysis, possible residual confounding by
unmeasured factors such as income or severity of hypertension, and the
clustering of patients within six facilities. Finally, the Discussion
should avoid the Table 2 fallacy: the adjusted ORs are mutually adjusted
associations, not a ranking of causal effects [@westreich2013].

## Reproducible reporting {#sec-reproducible}

### Why reproducibility matters

An analysis is *reproducible* if another person, given the same data and
code, obtains the same results [@peng2011]. This is a lower bar than
*replicability* (obtaining consistent results in new data), but it is
surprisingly often not met. Numbers copied by hand from the console into a
manuscript drift out of date as the analysis evolves; figures produced by
clicking through menus cannot be regenerated; and a script that runs only
on the author's computer, with files in `C:/Users/...`, cannot be checked by
anyone else. In clinical research, where results may change practice, the
ability to retrace every step from raw data to published number is part of
scientific integrity.

### Principles for a reproducible analysis

@wilson2017 describe "good enough" practices that any researcher can adopt,
and @peng2011 describes a spectrum from publication alone to fully linked
and executable code and data. For an analysis like the one in this book,
the essential practices are as follows.

- **Keep raw data read-only.** Never edit the raw file by hand. All
  cleaning is done by a script (Chapter 2) that writes a separate clean
  file, as `analysis_data.rds` was created.
- **Organise the project.** Use one folder per project, opened as an
  RStudio Project, with sub-folders such as `Data/` (raw and clean data),
  `Scripts/` (numbered analysis scripts), `Resources/` or `Output/`
  (generated tables and figures), and `Reports/` (R Markdown or Quarto
  documents).
- **Use relative paths.** Refer to files relative to the project folder
  (`"Data/analysis_data.rds"`), never with an absolute path; the
  `here` package helps when scripts live in sub-folders.
- **Script everything.** Every table and figure should be produced by code
  that is saved, not by interactive commands typed and forgotten.
- **Set the random seed.** Any analysis involving random numbers
  (bootstrapping, multiple imputation, random splits, simulation) must call
  `set.seed()` so that it gives identical results each time.
- **Record the software environment.** Report the versions of R and of
  the key packages, since results and defaults can change between
  versions.
- **Use version control** (such as Git) to keep a history of changes to
  code and text.

### Setting the seed: an example

Maximum likelihood estimation is deterministic, so `model_full` gives the
same coefficients every time. Resampling methods are not. A bootstrap
confidence interval for the AUC, for example, draws random resamples of
the patients.

```{r c5-seed}
set.seed(2026)
ci.auc(roc_obj, method = "bootstrap", boot.n = 500, progress = "none")

set.seed(2026)          # same seed, so exactly the same interval
ci.auc(roc_obj, method = "bootstrap", boot.n = 500, progress = "none")
```

Run with the same seed, the two bootstrap intervals are identical; with a
different seed (or none) they would differ slightly from run to run. Both
are close to the analytical DeLong interval computed earlier.

### From script to report: R Markdown and Quarto

The most powerful tool for reproducible reporting is the *dynamic
document*, which interleaves text and code in a single file
[@xie2015]. In an R Markdown (`.Rmd`) or Quarto (`.qmd`) document, prose
is written in Markdown and analysis code is placed in *chunks*. When the
document is rendered ("knitted"), R runs every chunk in a fresh session
and inserts the output (tables, figures and numbers) into a Word, PDF or
HTML report. This book itself was produced in this way.

A minimal analysis report has three parts:

1. A *header* (written in YAML) with the title, author, date and output
   format, for example `output: word_document` in R Markdown or
   `format: docx` in Quarto.
2. *Code chunks* that load the packages and data, fit the models and
   produce the tables and figures exactly as in this chapter.
3. *Text* with inline R code, so that numbers in sentences are computed
   rather than typed. The Results paragraph in Section 5.11 was written
   this way: an inline expression such as `nrow(dx)`, placed between
   backticks after the letter `r`, is replaced by its value
   (`r nrow(dx)`) when the document is rendered.

Because the report is regenerated from the data each time, a correction to
the cleaning script, or an updated data set, flows automatically into every
number, table and figure. There is no step at which a number can be copied
wrongly.

The code below shows how outputs would be saved from a script. In a real
project they would go to an output folder; here they are written to a
temporary directory so that nothing in the project is changed.

```{r c5-save-outputs}
out_dir <- tempdir()     # in a project: "Resources" or "Output"

# Regression table and forest plot data as CSV
write_csv(or_table, file.path(out_dir, "regression_table.csv"))

# The forest plot as a high-resolution PNG
forest_plot <- ggplot(forest_df, aes(x = estimate, y = label)) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.25,
                orientation = "y") +
  geom_point() +
  scale_x_log10()
ggsave(file.path(out_dir, "forest_plot.png"), forest_plot,
       width = 7, height = 5, dpi = 300)

list.files(out_dir, pattern = "regression_table|forest_plot")
```

### Recording the software environment

`sessionInfo()` prints the R version, the operating system and every loaded
package with its version. In a report, you would include it in an appendix
or supplementary file, for example with
`writeLines(capture.output(sessionInfo()), "session_info.txt")`. The full
output is long; the essential information is shown below.

```{r c5-session}
R.version.string
pkgs <- c("tidyverse", "broom", "gtsummary", "car", "pROC", "ggplot2")
tibble(package = pkgs,
       version = map_chr(pkgs, \(p) as.character(packageVersion(p))))
```

::: {.callout-tip title="Good practice"}
Before submitting a paper, restart R, clear the workspace, and run (or
render) the whole analysis from the raw data to the final report in one go.
If it runs without error and reproduces every number in the manuscript,
your analysis is reproducible. If it does not, it is better that you find
out now than a reviewer later.
:::

## Summary {#sec-summary-ch5}

This chapter brought together the skills of the previous chapters to answer
the study's main question with regression models. Linear regression models
the mean of a continuous outcome as a linear function of several
predictors, estimated by least squares; its coefficients are differences in
means, holding other predictors fixed, and its assumptions are checked with
residual plots. For a binary outcome, logistic regression models the
log-odds as a linear function of the predictors, estimated by maximum
likelihood; exponentiated coefficients are odds ratios, which approximate
risk ratios only when the outcome is rare.

We fitted a pre-specified multivariable logistic regression for treatment
uptake among 992 patients with diagnosed hypertension and complete data,
read the `summary()` output line by line, converted coefficients to odds
ratios with profile-likelihood confidence intervals, rescaled continuous
predictors to meaningful increments, and tested education as a whole with a
likelihood-ratio test. Comparing crude and adjusted estimates showed that
the apparent association between distance and uptake was largely explained
by residence, a clear example of confounding, while the changes in the ORs
for family history and insurance partly reflected non-collapsibility. The model
had adequate events per variable, no collinearity, no evidence of
non-linearity or poor calibration, and acceptable discrimination. Finally,
we presented the results as a merged crude and adjusted table and a forest
plot, wrote Methods and Results paragraphs in line with STROBE, and saw how
dynamic documents, seeds and a recorded software environment make the whole
analysis reproducible.

::: {.callout-important title="Key points"}
- Decide the purpose of a model (description, prediction or explanation)
  and its predictors before looking at the outcome data.
- In logistic regression, $\exp(\beta)$ is the odds ratio for a one-unit
  increase, holding the other predictors fixed; report continuous
  predictors per clinically meaningful increment, $\exp(c\beta)$.
- Odds ratios exaggerate risk ratios when the outcome is common; say "odds",
  and add predicted probabilities or risk ratios when readers need risks.
- Report the number of patients analysed: `glm()` silently drops anyone
  with a missing value in any model variable.
- Use likelihood-ratio tests (`drop1()`, `anova(test = "LRT")`) for
  multi-level factors and for comparing nested models fitted to the same
  patients.
- A change from crude to adjusted OR may reflect confounding, but ORs are
  non-collapsible; reason causally, and do not interpret every adjusted OR
  as a causal effect (the Table 2 fallacy).
- Avoid stepwise selection in explanatory models; check events per
  variable, collinearity (VIF), linearity, discrimination (AUC),
  calibration and influential observations.
- Present crude and adjusted ORs with 95% CIs in a `gtsummary` table and a
  log-scale forest plot; write results as associations, following STROBE.
- Make the analysis reproducible: scripts, relative paths, `set.seed()`,
  dynamic documents and a record of software versions.
:::

## Further reading {#sec-further-reading-ch5}

- @hosmer2013, *Applied Logistic Regression*: the standard reference on
  logistic regression, covering model building, interpretation of odds
  ratios, goodness of fit and diagnostics in depth.
- @vittinghoff2012, *Regression Methods in Biostatistics*: a clear,
  example-driven account of linear, logistic, Poisson and Cox regression,
  with an excellent chapter on choosing predictors for different purposes.
- @harrell2015, *Regression Modeling Strategies*: the definitive guide to
  building prediction models responsibly, including the case against
  stepwise selection, splines, validation and calibration.
- @westreich2013 and @vanderweele2019: two short, readable papers on how to
  choose adjustment variables and how (not) to interpret a multivariable
  model's coefficients.
- @vonelm2007, the STROBE statement: the checklist to follow when writing up
  any observational study.

## Exercises {#sec-exercises-ch5}

::: {.exercise title="Exercise 5.1"}
Using the full `analysis_data` (all 1,500 adults), fit a linear regression
of systolic blood pressure (`sbp_mmhg`) on `age`, `bmi` and `sex`.

1. Report the coefficient for age per 10 years, with its 95% CI.
2. Interpret the coefficient for `sexMale` in one sentence.
3. How much of the variation in SBP does the model explain?
4. Draw the residuals-versus-fitted plot. Are the assumptions reasonable?
:::

::: {.exercise title="Exercise 5.2"}
Among the diagnosed patients (`dx`), cross-tabulate `family_history_htn`
against `treatment_uptake`.

1. Calculate by hand the proportion and the odds of uptake in each group,
   and the crude odds ratio.
2. Fit `glm(treatment_uptake ~ family_history_htn, family = binomial)` and
   confirm that `exp(coef())` gives the same OR.
3. Calculate the crude risk ratio and compare it with the OR. Why do they
   differ? Check your answer with the Zhang and Yu formula.
:::

::: {.exercise title="Exercise 5.3"}
From `model_full`, calculate the adjusted odds ratio and 95% CI for:

1. a 10-year increase in age;
2. a 5-point increase in knowledge score;
3. a 10 km increase in distance to the facility.

Which of these increments do you consider most clinically meaningful, and
why should the increment be chosen before the analysis?
:::

::: {.exercise title="Exercise 5.4"}
Investigate whether the association between health insurance and uptake is
confounded.

1. Using the complete-case data `cc`, fit the crude model for
   `health_insurance` and compare its OR with the adjusted OR from
   `model_full`. Calculate the percentage change in the log OR.
2. Describe in words a causal diagram for insurance and uptake. Which of the
   other model variables might be confounders, and could any be mediators?
3. Is the change you observed best explained by confounding or by
   non-collapsibility? Justify your answer.
:::

::: {.exercise title="Exercise 5.5"}
Education is an ordered variable.

1. Use `drop1()` to obtain the likelihood-ratio test for education in
   `model_full`, and state the null hypothesis it tests.
2. Create a numeric version of education (0 = None, 1 = Primary,
   2 = Secondary, 3 = Tertiary) and refit the model with it in place of the
   factor. Interpret the OR "per level of education".
3. Compare the two models with AIC. Which would you report, and why?
:::

::: {.exercise title="Exercise 5.6"}
Patients were recruited from six facilities.

1. Add `facility` to `model_full` and test it with a likelihood-ratio test.
2. Does adding facility change any of the other adjusted ORs by more than
   10%?
3. Compare the AUC of the two models. What do you conclude about the role
   of facility in this analysis?
:::

::: {.exercise title="Exercise 5.7"}
A health economist suggests that health insurance might matter more for
rural patients, who face higher costs of travel.

1. Add the interaction `health_insurance:residence` to `model_full` and test
   it with a likelihood-ratio test.
2. From the interaction model, calculate the OR for insurance separately in
   rural and in urban patients.
3. Write two sentences reporting the result, as you would in a paper.
:::

::: {.exercise title="Exercise 5.8"}
A secondary outcome is blood pressure control (`bp_controlled`), defined
only among patients on treatment.

1. Restrict the data to diagnosed patients with `treatment_uptake == "Yes"`
   and tabulate `bp_controlled`. How many events are there?
2. Using the 10 events-per-variable guideline, how many parameters can a
   logistic model for BP control support?
3. Pre-specify and fit a model with no more than that number of parameters
   (for example age, sex, BMI and adherence). Report the adjusted ORs with
   95% CIs and the AUC.
4. Write a short Results paragraph (about 100 words) that reports the
   findings appropriately, including the limitations imposed by the small
   number of events.
:::
