## Chapter 3: Descriptive Statistics, Tables and Figures

```{r s3-setup, include=FALSE}
library(tidyverse)
library(gtsummary)
library(janitor)
library(knitr)
library(scales)
```

All solutions start from the clean analysis file and, where the outcome is involved, from
the population of diagnosed patients.

```{r s3-data}
analysis_data <- readRDS("Data/analysis_data.rds")
diagnosed <- analysis_data |> filter(htn_diagnosed == "Yes")
teal <- "#0D7377"
```

### Exercise 3.1

We compute the four statistics for each variable, together with the number of missing
values.

```{r s3-ex1}
analysis_data |>
  select(age, distance_to_facility_km) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  group_by(variable) |>
  summarise(
    missing = sum(is.na(value)),
    mean    = mean(value, na.rm = TRUE),
    sd      = sd(value, na.rm = TRUE),
    median  = median(value, na.rm = TRUE),
    iqr     = IQR(value, na.rm = TRUE)
  ) |>
  kable(digits = 2, caption = "Summaries of age and distance to the facility.")
```

Age has 2 missing values and distance 60. For **age**, the mean (52.3 years) and median (52
years) almost coincide and the SD (14.1) is small relative to the mean, so the distribution
is roughly symmetric: report **mean (SD)**, 52.3 (14.1) years. For **distance**, the mean
(7.8 km) is well above the median (6.45 km) and the SD (6.1 km) is close to the mean of a
variable that cannot be negative, the signature of right skew (see the histogram in Section
3.5): report **median (IQR)**: 6.45 km (IQR 4.0 to 10.3 km, an interquartile range of 6.3 km).

### Exercise 3.2

```{r s3-ex2}
ldl <- analysis_data$ldl_mmol_l

mean(ldl)                                   # the na.rm trap: NA

c(n_valid   = sum(!is.na(ldl)),
  n_missing = sum(is.na(ldl)),
  pct_miss  = 100 * mean(is.na(ldl)),
  mean      = mean(ldl, na.rm = TRUE),
  sd        = sd(ldl, na.rm = TRUE),
  cv_pct    = 100 * sd(ldl, na.rm = TRUE) / mean(ldl, na.rm = TRUE)) |>
  round(2)
```

Without `na.rm = TRUE`, `mean()` returns `NA` because 105 LDL values (7.0%) are missing.
Among the 1,395 patients with a result, mean LDL cholesterol was 2.70 mmol/L with SD 1.03
mmol/L, a coefficient of variation of about 38%, so LDL is relatively more variable than
total cholesterol (CV about 19%). A sentence for a paper: *"LDL cholesterol was available for
1,395 of 1,500 participants (93.0%); mean 2.70 mmol/L (SD 1.03)."*

### Exercise 3.3

```{r s3-ex3-table}
table(analysis_data$smoking, useNA = "ifany")
table(analysis_data$bp_category, useNA = "ifany")
```

```{r s3-ex3-tabyl}
analysis_data |>
  tabyl(smoking) |>
  adorn_pct_formatting(digits = 1)

analysis_data |>
  tabyl(bp_category) |>
  adorn_pct_formatting(digits = 1)
```

Smoking status is missing for 45 patients and blood-pressure category for 3. There are 261
current smokers: (a) 17.4% of all 1,500 patients (`percent`), and (b) 17.9% of the 1,455 with
known smoking status (`valid_percent`). The difference is small here because only 3% are
missing, but the denominator should always be stated.

### Exercise 3.4

```{r s3-ex4}
ins_tab <- table(Insurance = diagnosed$health_insurance,
                 Treated   = diagnosed$treatment_uptake)
addmargins(ins_tab)

round(100 * prop.table(ins_tab, margin = 1), 1)   # row % (within insurance)
round(100 * prop.table(ins_tab, margin = 2), 1)   # column % (within uptake)
```

Among the 364 insured diagnosed patients, 55.5% were on treatment, compared with 42.2% of the
725 uninsured. These are the **row percentages**, computed within categories of the
explanatory variable (insurance). They answer the question "is insurance associated with
uptake?" directly, because they compare the outcome between the exposure groups. The column
percentages (39.8% of treated and 27.9% of untreated patients were insured) describe the
make-up of the outcome groups, which is what a Table 1 by treatment uptake shows, but they do
not state the uptake rate in each insurance group.

### Exercise 3.5

```{r s3-ex5-mean}
diagnosed |>
  group_by(treatment_uptake) |>
  summarise(
    n     = sum(!is.na(sbp_mmhg)),
    mean  = mean(sbp_mmhg, na.rm = TRUE),
    sd    = sd(sbp_mmhg, na.rm = TRUE),
    se    = sd / sqrt(n),
    lower = mean - qt(0.975, n - 1) * se,
    upper = mean + qt(0.975, n - 1) * se
  ) |>
  kable(digits = 2, caption = "Mean SBP (mmHg) with 95% CI by uptake.")
```

```{r s3-ex5-prop}
diagnosed |>
  group_by(residence) |>
  summarise(n = n(), treated = sum(treatment_uptake == "Yes")) |>
  mutate(pct   = 100 * treated / n,
         se    = sqrt((pct / 100) * (1 - pct / 100) / n),
         lower = pct - 100 * 1.96 * se,
         upper = pct + 100 * 1.96 * se) |>
  select(-se) |>
  kable(digits = 1, caption = "Treatment uptake (%) with 95% CI by residence.")
```

Untreated diagnosed patients had mean SBP 142.2 mmHg (95% CI about 140.6 to 143.8) and treated
patients 148.1 mmHg (about 146.5 to 149.7). Uptake was 38.9% (95% CI about 34.7% to 43.2%)
among rural and 53.2% (about 49.2% to 57.3%) among urban diagnosed patients; the intervals do
not overlap, suggesting a real difference that Chapter 4 will test formally.

The **SD** (about 18 to 20 mmHg) measures how much individual patients' SBP varies around
their group mean; it would stay about the same if we recruited more patients. The **SE**
(about 0.8 mmHg) measures how precisely each group *mean* is estimated; it is the SD divided
by $\sqrt{n}$ and would shrink with a larger sample. The SD describes patients, the SE
describes the uncertainty of an estimate.

### Exercise 3.6

```{r s3-ex6}
diagnosed |>
  select(age, sex, residence, education, health_insurance,
         family_history_htn, knowledge_score, comorbidity_count,
         treatment_uptake) |>
  tbl_summary(
    by = treatment_uptake,
    type = comorbidity_count ~ "continuous",   # treat the count as numeric
    statistic = list(
      c(age, knowledge_score) ~ "{mean} ({sd})",
      comorbidity_count ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(c(age, knowledge_score) ~ 1, comorbidity_count ~ 0),
    label = list(
      age ~ "Age, years",
      sex ~ "Sex",
      residence ~ "Residence",
      education ~ "Education",
      health_insurance ~ "Health insurance",
      family_history_htn ~ "Family history of hypertension",
      knowledge_score ~ "Knowledge score (0-20)",
      comorbidity_count ~ "Number of comorbidities"
    ),
    missing = "ifany",
    missing_text = "Missing"
  ) |>
  add_overall(last = FALSE) |>
  modify_header(label = "**Characteristic**") |>
  as_kable(caption = "Diagnosed patients by treatment uptake.")
```

Without `type = comorbidity_count ~ "continuous"`, gtsummary would treat a count with only
four distinct values as categorical and show n (%) for each value, which is in fact a
perfectly good alternative here: the median (IQR) of 0 (0, 1) is identical in both groups and
hides any difference.

Two characteristics that differ: treated patients were **older** (mean about 56 versus 51
years) and more often had a **family history of hypertension** (about 42% versus 31%); they
were also more often insured, urban and better educated, and had a slightly higher knowledge
score. These differences are clinically plausible: older patients and those with a family
history are more likely to be recognised as high-risk and started on treatment, and
insurance, education and knowledge plausibly lower the financial and informational barriers
to care. They are unadjusted, descriptive differences in simulated data, to be examined with
tests (Chapter 4) and multivariable models (Chapter 5).

### Exercise 3.7

```{r s3-ex7a, fig.cap="Hypertension knowledge score (0-20) among diagnosed patients not on treatment and on treatment; boxes show the median and quartiles, points show individual patients."}
ggplot(diagnosed, aes(x = treatment_uptake, y = knowledge_score)) +
  geom_jitter(width = 0.2, height = 0.2, alpha = 0.25, size = 0.8,
              na.rm = TRUE) +
  geom_boxplot(width = 0.4, outlier.shape = NA, fill = teal, alpha = 0.3,
               na.rm = TRUE) +
  labs(title = "Treated patients score slightly higher on knowledge",
       x = "On antihypertensive treatment", y = "Knowledge score (0-20)")
```

The knowledge score is an integer, so a small vertical jitter (`height = 0.2`) stops patients
with the same score from being drawn on top of each other. The median score is 11 among
treated and 10 among untreated patients, but the two distributions overlap almost entirely.

```{r s3-ex7b, fig.cap="Percentage of diagnosed hypertensive patients on treatment by highest education level (24 patients with missing education excluded); uptake rises steadily with education."}
diagnosed |>
  filter(!is.na(education)) |>
  group_by(education) |>
  summarise(pct = mean(treatment_uptake == "Yes")) |>
  ggplot(aes(x = education, y = pct)) +
  geom_col(fill = teal, width = 0.65) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)), vjust = -0.4) +
  scale_y_continuous(labels = label_percent(), limits = c(0, 0.7)) +
  labs(title = "Treatment uptake rises with education",
       x = "Highest education level", y = "Diagnosed patients on treatment")
```

Because `education` is an ordered factor, the bars appear from none to tertiary without extra
code. Uptake rises from about 37% among patients with no schooling to about 57% among those
with tertiary education, a gradient consistent with education easing access to and use of
care. A bar chart is appropriate here because each bar is a *percentage*, whose length from
zero is meaningful.

### Exercise 3.8

```{r s3-ex8-table}
facility_summary <- diagnosed |>
  group_by(facility) |>
  summarise(
    n           = n(),
    treated     = sum(treatment_uptake == "Yes"),
    median_dist = median(distance_to_facility_km, na.rm = TRUE),
    pct_insured = 100 * mean(health_insurance == "Yes")
  ) |>
  rowwise() |>                                   # one prop.test() per facility
  mutate(
    ci    = list(prop.test(treated, n, correct = FALSE)$conf.int),
    pct   = 100 * treated / n,
    lower = 100 * ci[1],
    upper = 100 * ci[2]
  ) |>
  ungroup() |>
  select(facility, n, pct, lower, upper, median_dist, pct_insured) |>
  arrange(desc(pct))

facility_summary |>
  kable(digits = 1,
        caption = "Uptake (%, Wilson 95% CI) and access by facility.")
```

```{r s3-ex8-fig, fig.cap="Treatment uptake among diagnosed patients against median distance to the facility, one point per facility; with only six facilities no clear relationship is visible."}
ggplot(facility_summary, aes(x = median_dist, y = pct / 100)) +
  geom_point(size = 3, colour = teal) +
  geom_text(aes(label = facility), vjust = -1, size = 3.2) +
  scale_y_continuous(labels = label_percent(), limits = c(0.35, 0.58)) +
  scale_x_continuous(limits = c(5.2, 7.5)) +
  labs(title = "Facility uptake and median distance",
       x = "Median distance to facility (km)",
       y = "Diagnosed patients on treatment")
```

`rowwise()` makes `mutate()` work one facility at a time, so `prop.test()` receives that
facility's counts; the Wilson interval is stored as a list column and then split into its
lower and upper limits. Uptake ranges from 38.5% (Igoma HC) to 53.6% (Nyamagana PHC), with
Wilson intervals similar to the Wald intervals of Section 3.8 because each facility has more
than 100 diagnosed patients. In the plot there is no clear pattern: the facilities with the
shortest median distance (Buzuruga PHC and Igoma HC) do not have the highest uptake.

Great caution is needed for three reasons. First, there are only **six** data points, so any
apparent relationship could easily arise by chance. Second, this is an **ecological**
comparison of facility averages: a relationship between facilities need not hold for
individual patients (the *ecological fallacy*), and indeed at the patient level Table 1 in
Section 3.10 showed that treated patients lived somewhat closer to the facility. Third,
facilities differ in many ways at once (insurance coverage, staffing, catchment population),
so distance cannot be isolated from other facility characteristics without a
patient-level multivariable analysis that accounts for clustering (Chapter 5).
